---
title: "8. Visualising and Analysing Time-oriented Data"
author: "carina peh"
date: "June 11, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Time-oriented data captures how things change over time. In this exercise, three key visualisations for time-series analysis are explored: **calendar heatmaps** to spot attack patterns by hour and day, **cycle plots** to reveal seasonal trends in visitor arrivals, and **slopegraphs** to compare changes in rankings across two time points.

## Reflections

Time-series visualisations go far beyond a simple line chart. A calendar heatmap uses colour intensity to reveal *when* events cluster, for example, showing that cyber attacks spike on certain hours or weekdays. A cycle plot is particularly useful for seasonal data because it overlays multiple years on the same month panels, making trends immediately visible. A slopegraph is elegant for comparing two time points across many categories — the slope direction and angle tell the story at a glance. The key takeaway is that choosing the right chart type for time-oriented data depends on whether you are looking for *patterns within a cycle*, *trends over time*, or *relative change between points*.

## Getting Started

### Loading packages

```{r}
pacman::p_load(scales, viridis, lubridate, ggthemes,
               gridExtra, readxl, knitr, data.table,
               ggHoriPlot, tidyverse)
```

**What each package does:**

-   `scales` — formats axis labels (e.g. percentages, commas)
-   `viridis` — colour-blind friendly colour palettes
-   `lubridate` — makes working with dates and times easy
-   `ggthemes` — adds extra themes like `theme_tufte()`
-   `gridExtra` — arranges multiple plots side by side
-   `readxl` — reads Excel files (.xlsx)
-   `knitr` — formats tables nicely with `kable()`
-   `data.table` — fast data manipulation
-   `CGPfunctions` — provides `newggslopegraph()`
-   `ggHoriPlot` — for horizon charts
-   `tidyverse` — core suite: `ggplot2`, `dplyr`, `tidyr`, etc.

## Plotting Calendar Heatmap

A calendar heatmap plots a grid of **weekday × hour**, coloured by the count of events. It reveals *when* attacks are most frequent at a glance.

### The Data

*eventlog.csv* contains 199,999 rows of time-series cyber attack records. Each row records a timestamp, the source country, and the timezone.

### Importing the data

```{r}
attacks <- read_csv("eventlog.csv")
```

### Examining the data structure

```{r}
kable(head(attacks))
```

There are three columns:

-   `timestamp` — date-time of the attack in POSIXct format
-   `source_country` — ISO 3166-1 alpha-2 country code (e.g. "CN" for China)
-   `tz` — timezone of the source IP address

### Data Preparation

#### Step 1: Deriving *weekday* and *hour* fields

A helper function `make_hr_wkday()` is written to take the timestamp, country, and timezone, and return a clean data table with the weekday and hour extracted correctly per timezone.

```{r}
make_hr_wkday <- function(ts, sc, tz) {
  real_times <- ymd_hms(ts,
                        tz = tz[1],
                        quiet = TRUE)
  dt <- data.table(source_country = sc,
                   wkday = weekdays(real_times),
                   hour = hour(real_times))
  return(dt)
}
```

-   `ymd_hms()` from **lubridate** parses the timestamp string into a proper date-time object, applying the correct timezone
-   `weekdays()` from base R extracts the day of the week as a character string
-   `hour()` from **lubridate** extracts just the hour (0–23)

#### Step 2: Deriving the attacks tibble

```{r}
wkday_levels <- c('Saturday', 'Friday',
                  'Thursday', 'Wednesday',
                  'Tuesday', 'Monday',
                  'Sunday')

attacks <- attacks %>%
  group_by(tz) %>%
  do(make_hr_wkday(.$timestamp,
                   .$source_country,
                   .$tz)) %>%
  ungroup() %>%
  mutate(wkday = factor(
    wkday, levels = wkday_levels),
    hour  = factor(
      hour, levels = 0:23))
```

Both `wkday` and `hour` are converted to **factors** with explicit level ordering. This ensures ggplot2 displays them in the correct sequence (Sunday at the bottom, Saturday at the top; hours 0–23 left to right).

```{r}
kable(head(attacks))
```

### Building the Calendar Heatmap

::: panel-tabset
## The plot

```{r}
#| echo: false
grouped <- attacks %>%
  count(wkday, hour) %>%
  ungroup() %>%
  na.omit()

ggplot(grouped,
       aes(hour,
           wkday,
           fill = n)) +
geom_tile(color = "white",
          size = 0.1) +
theme_tufte(base_family = "Helvetica") +
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue",
                    high = "dark blue") +
labs(x = NULL,
     y = NULL,
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6))
```

## The code chunk

```{r}
#| eval: false
grouped <- attacks %>%
  count(wkday, hour) %>%
  ungroup() %>%
  na.omit()

ggplot(grouped,
       aes(hour,
           wkday,
           fill = n)) +
geom_tile(color = "white",
          size = 0.1) +
theme_tufte(base_family = "Helvetica") +
coord_equal() +
scale_fill_gradient(name = "# of attacks",
                    low = "sky blue",
                    high = "dark blue") +
labs(x = NULL,
     y = NULL,
     title = "Attacks by weekday and time of day") +
theme(axis.ticks = element_blank(),
      plot.title = element_text(hjust = 0.5),
      legend.title = element_text(size = 8),
      legend.text = element_text(size = 6))
```
:::

**What to notice:**

-   `geom_tile()` draws one rectangle per hour-weekday combination, filled by attack count
-   `theme_tufte()` strips away gridlines and chart junk for a cleaner look
-   `coord_equal()` ensures each tile is a perfect square (1:1 aspect ratio)
-   `scale_fill_gradient()` maps count to a sky blue → dark blue colour scale — darker means more attacks

### Building Multiple Calendar Heatmaps

**Challenge:** Plot separate heatmaps for the top 4 countries with the most attacks.

#### Step 1: Deriving attack count by country

```{r}
attacks_by_country <- count(
  attacks, source_country) %>%
  mutate(percent = percent(n/sum(n))) %>%
  arrange(desc(n))
```

#### Step 2: Filtering to the top 4 countries

```{r}
top4 <- attacks_by_country$source_country[1:4]
top4_attacks <- attacks %>%
  filter(source_country %in% top4) %>%
  count(source_country, wkday, hour) %>%
  ungroup() %>%
  mutate(source_country = factor(
    source_country, levels = top4)) %>%
  na.omit()
```

#### Step 3: Plotting the multiple calendar heatmaps

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(top4_attacks,
       aes(hour,
           wkday,
           fill = n)) +
  geom_tile(color = "white",
          size = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue",
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL,
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))
```

## The code chunk

```{r}
#| eval: false
ggplot(top4_attacks,
       aes(hour,
           wkday,
           fill = n)) +
  geom_tile(color = "white",
          size = 0.1) +
  theme_tufte(base_family = "Helvetica") +
  coord_equal() +
  scale_fill_gradient(name = "# of attacks",
                    low = "sky blue",
                    high = "dark blue") +
  facet_wrap(~source_country, ncol = 2) +
  labs(x = NULL, y = NULL,
     title = "Attacks on top 4 countries by weekday and time of day") +
  theme(axis.ticks = element_blank(),
        axis.text.x = element_text(size = 7),
        plot.title = element_text(hjust = 0.5),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 6))
```
:::

`facet_wrap(~source_country, ncol = 2)` splits the plot into a 2×2 grid, one panel per country. This makes it easy to compare attack patterns side-by-side.

## Plotting Cycle Plot

A cycle plot shows **seasonal patterns** by faceting across months, drawing each year as a separate line within each month panel. A red dashed horizontal line marks the monthly average, making it easy to see whether recent years are above or below the long-run norm.

### Step 1: Data Import

```{r}
air <- read_excel("arrivals_by_air.xlsx")
```

### Step 2: Deriving month and year fields

```{r}
air$month <- factor(month(air$`Month-Year`),
                    levels=1:12,
                    labels=month.abb,
                    ordered=TRUE)
air$year <- year(ymd(air$`Month-Year`))
```

-   `month()` extracts the month number (1–12), then labelled with standard three-letter abbreviations (`month.abb`)
-   `year()` extracts the four-digit year
-   Both use `lubridate` functions

### Step 3: Extracting the target country

The analysis focuses on Vietnam, filtering to 2010 onwards.

```{r}
Vietnam <- air %>%
  select(`Vietnam`,
         month,
         year) %>%
  filter(year >= 2010)
```

### Step 4: Computing year average arrivals by month

```{r}
hline.data <- Vietnam %>%
  group_by(month) %>%
  summarise(avgvalue = mean(`Vietnam`))
```

This creates a small summary table with one average value per month — used to draw the red reference line.

### Step 5: Plotting the cycle plot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot() +
  geom_line(data=Vietnam,
            aes(x=year,
                y=`Vietnam`,
                group=month),
            colour="black") +
  geom_hline(aes(yintercept=avgvalue),
             data=hline.data,
             linetype=6,
             colour="red",
             size=0.5) +
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```

## The code chunk

```{r}
#| eval: false
ggplot() +
  geom_line(data=Vietnam,
            aes(x=year,
                y=`Vietnam`,
                group=month),
            colour="black") +
  geom_hline(aes(yintercept=avgvalue),
             data=hline.data,
             linetype=6,
             colour="red",
             size=0.5) +
  facet_grid(~month) +
  labs(axis.text.x = element_blank(),
       title = "Visitor arrivals from Vietnam by air, Jan 2010-Dec 2019") +
  xlab("") +
  ylab("No. of Visitors") +
  theme_tufte(base_family = "Helvetica")
```
:::

**What to notice:**

-   `geom_line()` draws one line per month panel, with year on the x-axis — each panel shows how that particular month has trended over the years
-   `geom_hline()` draws a dashed red line at the per-month average — lines above it indicate above-average years
-   `facet_grid(~month)` creates 12 side-by-side panels, one per month

## Plotting Slopegraph

A slopegraph compares values for multiple categories at **two time points**, using the slope of the connecting line to show whether each category increased, decreased, or stayed the same. It is ideal for ranking comparisons.

### Step 1: Data Import

```{r}
rice <- read_csv("rice.csv")
```

### Step 2: Plotting the slopegraph

::: panel-tabset
## The plot

```{r}
#| echo: false
rice %>%
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  ggplot(aes(x = Year, y = Yield, group = Country)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_point(colour = "steelblue", size = 2) +
  geom_text(data = . %>% filter(Year == 1961),
            aes(label = Country), hjust = 1.15, size = 3) +
  geom_text(data = . %>% filter(Year == 1980),
            aes(label = Country), hjust = -0.15, size = 3) +
  labs(title = "Rice Yield of Top 11 Asian Countries",
       subtitle = "1961-1980",
       caption = "Prepared by: Carina",
       x = NULL, y = "Yield") +
  theme_tufte(base_family = "Helvetica") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

## The code chunk

```{r}
#| echo: false
rice %>%
  mutate(Year = factor(Year)) %>%
  filter(Year %in% c(1961, 1980)) %>%
  ggplot(aes(x = Year, y = Yield, group = Country)) +
  geom_line(colour = "steelblue", linewidth = 0.8) +
  geom_point(colour = "steelblue", size = 2) +
  geom_text(data = . %>% filter(Year == 1961),
            aes(label = Country), hjust = 1.15, size = 3) +
  geom_text(data = . %>% filter(Year == 1980),
            aes(label = Country), hjust = -0.15, size = 3) +
  labs(title = "Rice Yield of Top 11 Asian Countries",
       subtitle = "1961-1980",
       caption = "Prepared by: Carina",
       x = NULL, y = "Yield") +
  theme_tufte(base_family = "Helvetica") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```
:::

**What to notice:**

-   `mutate(Year = factor(Year))` converts the year from a number to a categorical label, this is required by `newggslopegraph()` to treat the two years as discrete endpoints rather than a continuous axis
-   `filter(Year %in% c(1961, 1980))` keeps only the two comparison years
-   `newggslopegraph(Year, Yield, Country)` maps time to the x-axis, the measured value (rice yield) to the y-axis, and labels each line by country
-   Lines sloping **upward** indicate yield increased; lines sloping **downward** indicate a decline between 1961 and 1980
