---
title: "5e. Treemap Visualisation with R"
author: "carina peh"
date: "May 13, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Treemaps display hierarchical data using nested rectangles where the size of each rectangle represents a quantitative value. In this exercise I used Singapore private property transaction data from 2018.

## Reflections

Treemaps are effective when you need to show both hierarchy and proportion in one chart. The size encoding shows total units sold at a glance, while the colour encoding adds a second layer of information on median unit price. The `treemapify` approach using ggplot2 syntax was more intuitive since the layering logic is familiar. The data wrangling step using `group_by()` and `summarise()` before plotting was a good reminder that most real-world visualisation work starts with data preparation.

## Getting Started

### Loading packages

```{r}
pacman::p_load(treemap, treemapify, tidyverse)
```

### Importing data

```{r}
realis2018 <- read_csv("realis2018.csv")
```

## Data Wrangling

### Grouped summaries without the Pipe

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`, 
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped, 
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE), 
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

### Grouped summaries with the pipe

```{r}
realis2018_summarised <- realis2018 %>% 
  group_by(`Project Name`,`Planning Region`, 
           `Planning Area`, `Property Type`, 
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE), 
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

### Filtering resale condominiums

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

## Designing Treemap with treemap Package

### Designing a static treemap

::: panel-tabset
## The plot

```{r}
#| echo: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## The code chunk

```{r}
#| eval: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### Working with vColor and type arguments

::: panel-tabset
## The plot

```{r}
#| echo: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## The code chunk

```{r}
#| eval: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type = "value",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### The "value" type treemap

::: panel-tabset
## The plot

```{r}
#| echo: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## The code chunk

```{r}
#| eval: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="value",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### The "manual" type treemap

::: panel-tabset
## The plot

```{r}
#| echo: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## The code chunk

```{r}
#| eval: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="RdYlBu", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

The colour scheme above is confusing because `RdYlBu` is a diverging palette but all values are positive. A single-colour palette like `Blues` is more appropriate.

::: panel-tabset
## The plot

```{r}
#| echo: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## The code chunk

```{r}
#| eval: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues", 
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### Working with algorithm argument

::: panel-tabset
## The plot

```{r}
#| echo: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues",
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## The code chunk

```{r}
#| eval: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues",
        algorithm = "squarified",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

### Using sortID

::: panel-tabset
## The plot

```{r}
#| echo: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues",
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## The code chunk

```{r}
#| eval: false
treemap(realis2018_selected,
        index=c("Planning Region", "Planning Area", "Project Name"),
        vSize="Total Unit Sold",
        vColor="Median Unit Price ($ psm)",
        type="manual",
        palette="Blues",
        algorithm = "pivotSize",
        sortID = "Median Transacted Price",
        title="Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```
:::

## Designing Treemap using treemapify Package

### Designing a basic treemap

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") +
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

## The code chunk

```{r}
#| eval: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start = "bottomleft") +
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```
:::

### Defining hierarchy

**Group by Planning Region**

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") +
  geom_treemap()
```

## The code chunk

```{r}
#| eval: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") +
  geom_treemap()
```
:::

**Group by Planning Area**

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap()
```

## The code chunk

```{r}
#| eval: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap()
```
:::

**Adding boundary lines**

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

## The code chunk

```{r}
#| eval: false
ggplot(data=realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```
:::

### With hierarchy and labels

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           label = `Project Name`,
           subgroup = `Planning Region`)) +
  geom_treemap() +
  geom_treemap_subgroup_border(colour = "grey40", size = 2) +
  geom_treemap_subgroup_text(place = "centre",
                             grow = FALSE,
                             alpha = 0.5,
                             colour = "black",
                             fontface = "italic",
                             min.size = 0) +
  geom_treemap_text(colour = "white",
                    place = "topleft",
                    reflow = TRUE) +
  scale_fill_gradient(low = "light yellow", high = "red") +
  theme(legend.position = "bottom")
```

## The code chunk

```{r}
#| eval: false
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`,
           label = `Project Name`,
           subgroup = `Planning Region`)) +
  geom_treemap() +
  geom_treemap_subgroup_border(colour = "grey40", size = 2) +
  geom_treemap_subgroup_text(place = "centre",
                             grow = FALSE,
                             alpha = 0.5,
                             colour = "black",
                             fontface = "italic",
                             min.size = 0) +
  geom_treemap_text(colour = "white",
                    place = "topleft",
                    reflow = TRUE) +
  scale_fill_gradient(low = "light yellow", high = "red") +
  theme(legend.position = "bottom")
```
:::

## Designing Interactive Treemap using d3treeR

### Installing d3treeR

```{r}
#| eval: false

install.packages("devtools")
install.packages("treemap")

library(devtools)
install_github("timelyportfolio/d3treeR")
```

### Loading packages

```{r}
library(treemap)
library(d3treeR)
```

### Designing an interactive treemap

```{r}
tm <- treemap(
  realis2018_summarised,
  index = c("Planning Region", "Planning Area"),
  vSize = "Total Unit Sold",
  vColor = "Median Unit Price ($ psm)",
  type = "value",
  title = "Private Residential Property Sold, 2017",
  title.legend = "Median Unit Price (S$ per sq. m)"
)
```

```{r}
d3tree(tm, rootname = "Singapore")
```

