---
title: "Hands-on Exercise 4c: Visualising Uncertainty"
author: "carina peh"
date: "May 6, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Visualising uncertainty is about showing not just the point estimate (like a mean) but also how confident we are in that estimate. In this exercise I learnt to plot error bars, confidence intervals and uncertainty using `ggdist`.

## Reflections

The difference between standard error bars and confidence intervals was much clearer after seeing them plotted side by side. The 95% CI version (mean ± 1.96 × SE) is more informative than just showing SE.

The interactive version combining `plotly` and `DT` was the most practical, being able to hover over a point and see the exact confidence interval values, and then filter the table at the same time, is very useful for presentations.

`stat_gradientinterval()` from `ggdist` was visually striking — the gradient effect gives a more intuitive feel for where the uncertainty is concentrated.

## Getting Started

### Loading packages

```{r}
pacman::p_load(plotly, crosstalk, DT,
               ggdist, ggridges, colorspace,
               gganimate, tidyverse)
```

### Importing data

```{r}
exam <- read_csv("Exam_data.csv")
```

### Computing summary statistics

```{r}
my_sum <- exam %>%
  group_by(RACE) %>%
  summarise(
    n=n(),
    mean=mean(MATHS),
    sd=sd(MATHS)
    ) %>%
  mutate(se=sd/sqrt(n-1))
```

## Visualising Uncertainty — ggplot2 methods

### Standard error bars

::: panel-tabset

## The plot

```{r}
#| echo: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE,
        ymin=mean-se,
        ymax=mean+se),
    width=0.2,
    colour="black",
    alpha=0.9,
    linewidth=0.5) +
  geom_point(aes(x=RACE,
                 y=mean),
             stat="identity",
             color="red",
             size = 1.5,
             alpha=1) +
  ggtitle("Standard error of mean maths score by race")
```

## The code chunk

```{r}
#| eval: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=RACE,
        ymin=mean-se,
        ymax=mean+se),
    width=0.2,
    colour="black",
    alpha=0.9,
    linewidth=0.5) +
  geom_point(aes(x=RACE,
                 y=mean),
             stat="identity",
             color="red",
             size = 1.5,
             alpha=1) +
  ggtitle("Standard error of mean maths score by race")
```

:::

### 95% Confidence interval

::: panel-tabset

## The plot

```{r}
#| echo: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean),
        ymin=mean-1.96*se,
        ymax=mean+1.96*se),
    width=0.2,
    colour="black",
    alpha=0.9,
    linewidth=0.5) +
  geom_point(aes(x=RACE,
                 y=mean),
             stat="identity",
             color="red",
             size = 1.5,
             alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

## The code chunk

```{r}
#| eval: false
ggplot(my_sum) +
  geom_errorbar(
    aes(x=reorder(RACE, -mean),
        ymin=mean-1.96*se,
        ymax=mean+1.96*se),
    width=0.2,
    colour="black",
    alpha=0.9,
    linewidth=0.5) +
  geom_point(aes(x=RACE,
                 y=mean),
             stat="identity",
             color="red",
             size = 1.5,
             alpha=1) +
  labs(x = "Maths score",
       title = "95% confidence interval of mean maths score by race")
```

:::

### Interactive error bars — 99% CI

::: panel-tabset

## The plot

```{r}
#| echo: false
shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se,
                     ymax=mean+2.58*se),
                     width=0.2,
                     colour="black",
                     alpha=0.9,
                     size=0.5) +
                   geom_point(aes(
                     x=RACE,
                     y=mean,
                     text = paste("Race:", `RACE`,
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[",
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity",
                     color="red",
                     size = 1.5,
                     alpha=1) +
                   xlab("Race") +
                   ylab("Average Scores") +
                   theme_minimal() +
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average maths scores by race")),
                tooltip = "text"),
       DT::datatable(shared_df,
                     rownames = FALSE,
                     class="compact",
                     width="100%",
                     options = list(pageLength = 10,
                                    scrollX=T),
                     colnames = c("No. of pupils",
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```

## The code chunk

```{r}
#| eval: false
shared_df = SharedData$new(my_sum)

bscols(widths = c(4,8),
       ggplotly((ggplot(shared_df) +
                   geom_errorbar(aes(
                     x=reorder(RACE, -mean),
                     ymin=mean-2.58*se,
                     ymax=mean+2.58*se),
                     width=0.2,
                     colour="black",
                     alpha=0.9,
                     size=0.5) +
                   geom_point(aes(
                     x=RACE,
                     y=mean,
                     text = paste("Race:", `RACE`,
                                  "<br>N:", `n`,
                                  "<br>Avg. Scores:", round(mean, digits = 2),
                                  "<br>95% CI:[",
                                  round((mean-2.58*se), digits = 2), ",",
                                  round((mean+2.58*se), digits = 2),"]")),
                     stat="identity",
                     color="red",
                     size = 1.5,
                     alpha=1) +
                   xlab("Race") +
                   ylab("Average Scores") +
                   theme_minimal() +
                   theme(axis.text.x = element_text(
                     angle = 45, vjust = 0.5, hjust=1)) +
                   ggtitle("99% Confidence interval of average maths scores by race")),
                tooltip = "text"),
       DT::datatable(shared_df,
                     rownames = FALSE,
                     class="compact",
                     width="100%",
                     options = list(pageLength = 10,
                                    scrollX=T),
                     colnames = c("No. of pupils",
                                  "Avg Scores",
                                  "Std Dev",
                                  "Std Error")) %>%
         formatRound(columns=c('mean', 'sd', 'se'),
                     digits=2))
```

:::

## Visualising Uncertainty — ggdist methods

### stat_pointinterval()

::: panel-tabset

## The plot

```{r}
#| echo: false
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval() +
  labs(title = "Visualising confidence intervals of mean math score",
       subtitle = "Mean Point + Multiple-interval plot")
```

## The code chunk

```{r}
#| eval: false
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval() +
  labs(title = "Visualising confidence intervals of mean math score",
       subtitle = "Mean Point + Multiple-interval plot")
```

:::

### stat_pointinterval() with .width

::: panel-tabset

## The plot

```{r}
#| echo: false
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(.width = 0.95,
                     .point = median,
                     .interval = qi) +
  labs(title = "Visualising confidence intervals of mean math score",
       subtitle = "Mean Point + Multiple-interval plot")
```

## The code chunk

```{r}
#| eval: false
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_pointinterval(.width = 0.95,
                     .point = median,
                     .interval = qi) +
  labs(title = "Visualising confidence intervals of mean math score",
       subtitle = "Mean Point + Multiple-interval plot")
```

:::

### stat_gradientinterval()

::: panel-tabset

## The plot

```{r}
#| echo: false
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_gradientinterval(
    fill = "skyblue",
    show.legend = TRUE
  ) +
  labs(title = "Visualising confidence intervals of mean math score",
       subtitle = "Gradient + interval plot")
```

## The code chunk

```{r}
#| eval: false
exam %>%
  ggplot(aes(x = RACE,
             y = MATHS)) +
  stat_gradientinterval(
    fill = "skyblue",
    show.legend = TRUE
  ) +
  labs(title = "Visualising confidence intervals of mean math score",
       subtitle = "Gradient + interval plot")
```

:::


