---
title: "Hands-on Exercise 4a: Visualising Distribution"
author: "carina peh"
date: "May 6, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

In this exercise I learnt two relatively new statistical graphic methods for visualising distribution — ridgeline plots and raincloud plots — using `ggridges` and `ggdist`.

## Getting Started

### Loading packages

```{r}
pacman::p_load(ggdist, ggridges, ggthemes,
               colorspace, tidyverse)
```

### Importing data

```{r}
exam <- read_csv("Exam_data.csv")
```

## Visualising Distribution with Ridgeline Plot

A ridgeline plot (also called a Joyplot) reveals the distribution of a numeric value for several groups. It works best when there are 5 or more groups with a clear pattern or ranking.

### Basic ridgeline plot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## The code chunk

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#7097BB", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
    ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

### Varying fill colours along the x axis

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```

## The code chunk

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = stat(x))) +
  geom_density_ridges_gradient(
    scale = 3,
    rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Temp. [F]",
                       option = "C") +
  scale_x_continuous(
    name = "English grades",
    expand = c(0, 0)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_ridges()
```
:::

### Mapping probabilities directly onto colour

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient",
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```

## The code chunk

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = 0.5 - abs(0.5-stat(ecdf)))) +
  stat_density_ridges(geom = "density_ridges_gradient",
                      calc_ecdf = TRUE) +
  scale_fill_viridis_c(name = "Tail probability",
                       direction = -1) +
  theme_ridges()
```
:::

### Ridgeline plots with quantile lines

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```

## The code chunk

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = 4,
    quantile_lines = TRUE) +
  scale_fill_viridis_d(name = "Quartiles") +
  theme_ridges()
```
:::

### Ridgeline plot with tail probability cut points

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```

## The code chunk

```{r}
#| eval: false
ggplot(exam,
       aes(x = ENGLISH,
           y = CLASS,
           fill = factor(stat(quantile))
           )) +
  stat_density_ridges(
    geom = "density_ridges_gradient",
    calc_ecdf = TRUE,
    quantiles = c(0.025, 0.975)
    ) +
  scale_fill_manual(
    name = "Probability",
    values = c("#FF0000A0", "#A0A0A0A0", "#0000FFA0"),
    labels = c("(0, 0.025]", "(0.025, 0.975]", "(0.975, 1]")
  ) +
  theme_ridges()
```
:::

## Visualising Distribution with Raincloud Plot

A raincloud plot combines a half-density plot, a boxplot and a dot plot — giving a much richer picture of the data distribution than a boxplot alone.

### Step 1 — Half Eye graph

::: panel-tabset
## The plot

```{r}
#| echo: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```

## The code chunk

```{r}
#| eval: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA)
```
:::

### Step 2 — Adding boxplot with geom_boxplot()

::: panel-tabset
## The plot

```{r}
#| echo: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```

## The code chunk

```{r}
#| eval: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA)
```
:::

### Step 3 — Adding dot plots stat_dots()

::: panel-tabset
## The plot

```{r}
#| echo: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = .5,
            dotsize = 2)
```

## The code chunk

```{r}
#| eval: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = .5,
            dotsize = 2)
```
:::

### Step 4 — Final raincloud plot

::: panel-tabset
## The plot

```{r}
#| echo: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = .5,
            dotsize = 2) +
  coord_flip() +
  theme_economist()
```

## The code chunk

```{r}
#| eval: false
exam %>%
  ggplot(aes(x = RACE,
             y = ENGLISH)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA) +
  geom_boxplot(width = .20,
               outlier.shape = NA) +
  stat_dots(side = "left",
            justification = 1.2,
            binwidth = .5,
            dotsize = 2) +
  coord_flip() +
  theme_economist()
```
:::

## Reflections

This exercise introduced two new ways to visualise distribution. Ridgeline plots are great for comparing distributions across many groups at once, much cleaner than stacking multiple histograms.

The raincloud plot was the most interesting, it combines three views in one chart: the density shape, the boxplot summary, and the actual data points. This gives a much more complete picture than just a boxplot alone.

The step-by-step build up of the raincloud plot (half eye → boxplot → dots → flip) was a good way to understand how each layer adds information.

## References

-   [ggridges documentation](https://wilkelab.org/ggridges/)
-   [ggdist documentation](https://mjskay.github.io/ggdist/)
-   [R for Visual Analytics Chapter 9](https://r4va.netlify.app/chap09)
