---
title: "Hands-on Exercise 4d: Building Funnel Plot with R"
author: "carina peh"
date: "May 6, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

A funnel plot is used for fair comparison between groups of different sizes. In this exercise I used COVID-19 data from Jakarta to compare cumulative fatality rates across sub-districts.

## Reflections

Funnel plots were completely new to me. The key insight is that comparing raw rates between areas of different sizes is unfair, a small sub-district with 2 deaths out of 10 cases looks much worse than a large one with 200 deaths out of 5000 cases, even if the true rate is similar.

The funnel shape comes from the fact that uncertainty decreases as sample size grows, small areas sit further out in the funnel, and outliers outside the 95% or 99% limits are the ones worth investigating.

Building the ggplot2 version from scratch was challenging but helped me understand exactly how the confidence limits are calculated.

## Getting Started

### Loading packages

```{r}
pacman::p_load(tidyverse, FunnelPlotR, plotly, knitr)
```

### Importing data

```{r}
covid19 <- read_csv("COVID-19_DKI_Jakarta.csv") %>%
  mutate_if(is.character, as.factor)
```

## FunnelPlotR methods

### Basic funnel plot

::: panel-tabset
## The plot

```{r}
#| echo: false
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```

## The code chunk

```{r}
#| eval: false
funnel_plot(
  .data = covid19,
  numerator = Positive,
  denominator = Death,
  group = `Sub-district`
)
```
:::

### Makeover 1 — using proportions

::: panel-tabset
## The plot

```{r}
#| echo: false
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05)
)
```

## The code chunk

```{r}
#| eval: false
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05)
)
```
:::

### Makeover 2 — with proper labels

::: panel-tabset
## The plot

```{r}
#| echo: false
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",
  x_label = "Cumulative COVID-19 Positive Cases",
  y_label = "Cumulative Fatality Rate"
)
```

## The code chunk

```{r}
#| eval: false
funnel_plot(
  .data = covid19,
  numerator = Death,
  denominator = Positive,
  group = `Sub-district`,
  data_type = "PR",
  xrange = c(0, 6500),
  yrange = c(0, 0.05),
  label = NA,
  title = "Cumulative COVID-19 Fatality Rate by Cumulative Total Number of COVID-19 Positive Cases",
  x_label = "Cumulative COVID-19 Positive Cases",
  y_label = "Cumulative Fatality Rate"
)
```
:::

## Funnel Plot using ggplot2

### Computing derived fields

```{r}
df <- covid19 %>%
  mutate(rate = Death / Positive) %>%
  mutate(rate.se = sqrt((rate*(1-rate)) / (Positive))) %>%
  filter(rate > 0)

fit.mean <- weighted.mean(df$rate, 1/df$rate.se^2)
```

### Calculate 95% and 99.9% confidence limits

```{r}
number.seq <- seq(1, max(df$Positive), 1)
number.ll95 <- fit.mean - 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq))
number.ul95 <- fit.mean + 1.96 * sqrt((fit.mean*(1-fit.mean)) / (number.seq))
number.ll999 <- fit.mean - 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq))
number.ul999 <- fit.mean + 3.29 * sqrt((fit.mean*(1-fit.mean)) / (number.seq))
dfCI <- data.frame(number.ll95, number.ul95,
                   number.ll999, number.ul999,
                   number.seq, fit.mean)
```

### Static funnel plot

::: panel-tabset
## The plot

```{r}
#| echo: false
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label=`Sub-district`),
             alpha=0.4,
             size=0.9) +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ll95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ul95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ll999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ul999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_hline(data = dfCI,
             aes(yintercept = fit.mean),
             linewidth = 0.4,
             colour = "grey40") +
  coord_cartesian(ylim=c(0,0.05)) +
  annotate("text", x = 1, y = -0.13, label = "95%", size = 3, colour = "grey40") +
  annotate("text", x = 4.5, y = -0.18, label = "99%", size = 3, colour = "grey40") +
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") +
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12),
        legend.position = c(0.91,0.85),
        legend.title = element_text(size=7),
        legend.text = element_text(size=7),
        legend.background = element_rect(colour = "grey60", linetype = "dotted"),
        legend.key.height = unit(0.3, "cm"))
p
```

## The code chunk

```{r}
#| eval: false
p <- ggplot(df, aes(x = Positive, y = rate)) +
  geom_point(aes(label=`Sub-district`),
             alpha=0.4,
             size=0.9) +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ll95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ul95),
            linewidth = 0.4,
            colour = "grey40",
            linetype = "dashed") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ll999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_line(data = dfCI,
            aes(x = number.seq,
                y = number.ul999),
            linewidth = 0.4,
            colour = "grey40") +
  geom_hline(data = dfCI,
             aes(yintercept = fit.mean),
             linewidth = 0.4,
             colour = "grey40") +
  coord_cartesian(ylim=c(0,0.05)) +
  ggtitle("Cumulative Fatality Rate by Cumulative Number of COVID-19 Cases") +
  xlab("Cumulative Number of COVID-19 Cases") +
  ylab("Cumulative Fatality Rate") +
  theme_light() +
  theme(plot.title = element_text(size=12))
p
```
:::

### Interactive funnel plot

::: panel-tabset
## The plot

```{r}
#| echo: false
fp_ggplotly <- ggplotly(p,
  tooltip = c("label",
              "x",
              "y"))
fp_ggplotly
```

## The code chunk

```{r}
#| eval: false
fp_ggplotly <- ggplotly(p,
  tooltip = c("label",
              "x",
              "y"))
fp_ggplotly
```
:::
