---
title: "Hands-on Exercise 4b: Visual Statistical Analysis"
author: "carina peh"
date: "May 6, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Reflections

`ggstatsplot` is really useful because it combines the chart and the statistical test result in one — something that would normally take two separate steps in R. The `ggbetweenstats()` function was the most interesting since it automatically runs the right test depending on the number of groups and shows the pairwise comparisons directly on the chart.

The Bayes Factor in `gghistostats()` was new to me — it gives a more intuitive way to interpret evidence compared to just a p-value.

## Getting Started

### Loading packages

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

### Importing data

```{r}
exam <- read_csv("Exam_data.csv")
```

## One-sample test: gghistostats()

Testing whether the mean English score is significantly different from 60.

::: panel-tabset
## The plot

```{r}
#| echo: false
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

## The code chunk

```{r}
#| eval: false
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```
:::

## Two-sample mean test: ggbetweenstats()

Comparing Maths scores between genders using a non-parametric test.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

## The code chunk

```{r}
#| eval: false
ggbetweenstats(
  data = exam,
  x = GENDER,
  y = MATHS,
  type = "np",
  messages = FALSE
)
```
:::

## Oneway ANOVA Test: ggbetweenstats()

Comparing English scores across different races with pairwise comparisons.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggbetweenstats(
  data = exam,
  x = RACE,
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

## The code chunk

```{r}
#| eval: false
ggbetweenstats(
  data = exam,
  x = RACE,
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```
:::

## Significant Test of Correlation: ggscatterstats()

Testing whether there is a significant correlation between Maths and English scores.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE
)
```

## The code chunk

```{r}
#| eval: false
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE
)
```
:::

## Significant Test of Association: ggbarstats()

Testing the association between Maths results (pass/fail) and gender.

::: panel-tabset
## The plot

```{r}
#| echo: false
exam1 <- exam %>%
  mutate(MATHS_bins =
           cut(MATHS,
               breaks = c(0,60,75,85,100))
)

ggbarstats(exam1,
           x = MATHS_bins,
           y = GENDER)
```

## The code chunk

```{r}
#| eval: false
exam1 <- exam %>%
  mutate(MATHS_bins =
           cut(MATHS,
               breaks = c(0,60,75,85,100))
)

ggbarstats(exam1,
           x = MATHS_bins,
           y = GENDER)
```
:::

## 
