---
title: "Hands-on Exercise 1: A Layered Grammar of Graphics"
author: "carinapeh"
date: "April 10, 2026"
date-modified: "last-modified"
description: "Introduction to ggplot2 and building statistical charts in R."
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Reflections

In this exercise, I was introduced to ggplot2 and the layered grammar of graphics. I learned that every chart in R is built by adding layers — starting from data, then aesthetics, then geometry. I explored many different chart types including histograms, dot plots, box plots, violin plots, and scatter plots. I also discovered how facets allow us to split charts by categories, and how built-in themes like theme_classic() and theme_minimal() can instantly change the look of a chart. This exercise showed me that ggplot2 is both systematic and flexible.

------------------------------------------------------------------------

# Getting Started

## Install and launching R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

```{r}
exam_data <- read_csv("Exam_data.csv")
```

------------------------------------------------------------------------

# Plotting a Simple Bar Chart

## Loading Tidyverse

This code chunk uses pacman's [p_load()](https://cran.r-project.org/web/packages/pacman/refman/pacman.html#p_load) to load tidyverse family of packages into R environment.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

## The code

```{r}
#| eval: false
ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```
:::

------------------------------------------------------------------------

# Geometric Objects

## R Graphics histogram

::: panel-tabset
## The plot

```{r}
#| echo: false
hist(exam_data$MATHS)
```

## The code

```{r}
#| eval: false
hist(exam_data$MATHS)
```
:::

## ggplot2 histogram

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=10, 
                 boundary = 100,
                 color="black", 
                 fill="grey") +
  ggtitle("Distribution of Maths scores")
```
:::

## Empty canvas

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data)
```
:::

## Canvas with aesthetic

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS))
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS))
```
:::

## Dot plot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(dotsize = 0.5)
```
:::

## Dot plot with cleaned y-axis

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_dotplot(binwidth=2.5,         
               dotsize = 0.5) +      
  scale_y_continuous(NULL,           
                     breaks = NULL)
```
:::

## Default histogram

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram()
```
:::

## Histogram with 20 bins

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20,            
                 color="black",      
                 fill="light blue")
```
:::

## Histogram by gender

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           fill = GENDER)) +
  geom_histogram(bins=20, 
                 color="grey30")
```
:::

## Basic density plot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_density()
```
:::

## Density by gender

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x = MATHS, 
           colour = GENDER)) +
  geom_density()
```
:::

## Basic boxplot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(y = MATHS,       
           x= GENDER)) +    
  geom_boxplot()
```
:::

## Notched boxplot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot(notch=TRUE)
```
:::

## Violin plot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_violin()
```
:::

## Basic scatterplot

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point()
```
:::

## Boxplot with jitter points

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(y = MATHS, 
           x= GENDER)) +
  geom_boxplot() +                    
  geom_point(position="jitter", 
             size = 0.5)
```
:::

## Boxplot only

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot()
```
:::

## Boxplot with mean point

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(y = MATHS, x= GENDER)) +
  geom_boxplot() +
  stat_summary(geom = "point",       
               fun = "mean",         
               colour ="red",        
               size=4)
```
:::

------------------------------------------------------------------------

# Scatterplots and Smooth Lines

## Scatterplot with curved smooth line

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(linewidth=0.5)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(linewidth=0.5)
```
:::

## Scatterplot with linear regression line

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5)
```
:::

## Scatterplot with axis limits

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS, y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100))
```
:::

------------------------------------------------------------------------

# Facets

## Facet wrap by class

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_wrap(~ CLASS)
```
:::

## Facet grid by class

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x= MATHS)) +
  geom_histogram(bins=20) +
    facet_grid(~ CLASS)
```
:::

------------------------------------------------------------------------

# Coordinates and Themes

## Basic bar chart (vertical)

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar()
```
:::

## Flipped bar chart (horizontal)

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip()
```
:::

## Bar Chart using theme_gray

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_gray()
```
:::

## Bar Chart using theme_classic

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_classic()
```
:::

## Bar Chart using theme_minimal

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data, 
       aes(x=RACE)) +
  geom_bar() +
  coord_flip() +
  theme_minimal()
```
:::
