---
title: "Hands-on Exercise 3b: Programming Animated Statistical Graphics with R"
author: "carina peh"
date: "April 29, 2026"
date-modified: "last-modified"
description: "Learnt how to create animated bubble charts using gganimate and plotly to tell data stories across time."
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Animated graphics can make data stories more engaging and memorable compared to static charts. In this exercise, I learnt how to create animated visualisations using **gganimate** and **plotly**, while also practising data wrangling with **tidyr** and **dplyr**.

Before diving in, it's worth asking, does animation actually add value here? For exploratory analysis, probably not. But for presentations, a well-placed animation can really help the audience connect with the data.

## Reflections

This was my first time making charts interactive in R. The `ggiraph` package was surprisingly straightforward — just swap `geom_point()` for `geom_point_interactive()` and wrap it in `girafe()`.

The most interesting part was the coordinated multiple views where hovering on one chart highlights the same student in another chart. This is really useful for spotting patterns across different variables at the same time.

I found `plotly` easier to use than `ggiraph` since `ggplotly()` can just wrap an existing ggplot chart. The crosstalk + DT combination was also very cool, being able to filter a chart by clicking on a table row is something I can see being very useful in real dashboards.

## Getting Started

### Loading packages

```{r}
pacman::p_load(readxl, gifski, gapminder,
               plotly, gganimate, tidyverse)
```

### Importing data

This exercise uses a different dataset — `GlobalPopulation.xls` instead of `Exam_data.csv`.

```{r}
col <- c("Country", "Continent")
globalPop <- read_xls("GlobalPopulation.xls",
                      sheet="Data") %>%
  mutate(across(all_of(col), as.factor)) %>%
  mutate(Year = as.integer(Year))
```

## Animated Data Visualisation — gganimate

### Static bubble plot first

Always good to build the static version before animating — helps check the layout looks right.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young')
```

## The code chunk

```{r}
#| eval: false
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young')
```
:::

### Animated bubble plot

Adding `transition_time()` animates the chart across years. `ease_aes('linear')` makes the movement smooth and consistent.

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') +
  transition_time(Year) +       
  ease_aes('linear')
```

## The code chunk

```{r}
#| eval: false
ggplot(globalPop, aes(x = Old, y = Young, 
                      size = Population, 
                      colour = Country)) +
  geom_point(alpha = 0.7, 
             show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(title = 'Year: {frame_time}', 
       x = '% Aged', 
       y = '% Young') +
  transition_time(Year) +       
  ease_aes('linear')
```
:::

## Animated Data Visualisation — plotly

### Using ggplotly()

This converts the ggplot animation into an interactive plotly chart with a play/pause button and a year slider — much more user-friendly for presentations.

::: panel-tabset
## The plot

```{r}
#| echo: false
gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') + 
  theme(legend.position='none')

ggplotly(gg)
```

## The code chunk

```{r}
#| eval: false
gg <- ggplot(globalPop, 
       aes(x = Old, 
           y = Young, 
           size = Population, 
           colour = Country)) +
  geom_point(aes(size = Population,
                 frame = Year),
             alpha = 0.7) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  labs(x = '% Aged', 
       y = '% Young') + 
  theme(legend.position='none')

ggplotly(gg)
```
:::

### Using plot_ly()

Building the animated chart directly with `plot_ly()` — slightly different syntax but gives more control over the output.

::: panel-tabset
## The plot

```{r}
#| echo: false
bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers') %>%
  layout(showlegend = FALSE)
bp
```

## The code chunk

```{r}
#| eval: false
bp <- globalPop %>%
  plot_ly(x = ~Old, 
          y = ~Young, 
          size = ~Population, 
          color = ~Continent,
          sizes = c(2, 100),
          frame = ~Year, 
          text = ~Country, 
          hoverinfo = "text",
          type = 'scatter',
          mode = 'markers') %>%
  layout(showlegend = FALSE)
bp
```
:::
