---
title: "9c. Analytical Mapping"
description: "Analytical mapping goes beyond visualisation by incorporating statistical techniques — rate maps, percentile maps, and box maps — to surface spatial outliers and extreme values."
author: "carina peh"
date: "June 20, 2026"
date-modified: "last-modified"
execute:
  eval: false
  echo: true
  warning: false
  freeze: true
---

## Overview

Analytical mapping extends basic choropleth mapping by incorporating statistical methods to highlight patterns, rates and outliers. In this exercise, three analytical mapping techniques are explored using Nigeria water point data: rate maps that show proportions rather than raw counts, percentile maps that highlight the extremes of a distribution, and box maps that flag statistical outliers.

## Reflections

This exercise changed how I think about spatial data. Mapping raw counts of water points is misleading — areas with more water points simply have more because they are larger or more populated, not because they are better served. Switching to a **rate map** (proportion of functional water points) immediately produces a more honest picture. The **percentile map** was particularly powerful: isolating the bottom 1% and top 1% makes extreme values jump out visually in a way that standard classification methods conceal. Writing reusable R functions (`get.var()`, `boxbreaks()`, `percentmap()`, `boxmap()`) also reinforced the principle of not repeating code unnecessarily.

## Getting Started

### Loading Packages

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

### Importing Data

```{r}
# NGA_wp is a polygon sf data frame of Nigeria Local Government Areas
# with water point statistics attached
NGA_wp <- read_rds("data/rds/NGA_wp.rds")
```

## Basic Choropleth Maps

```{r}
# Side-by-side: total water points vs functional water points
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
              fill.scale = tm_scale_intervals(style = "equal", n = 10,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of functional water points by LGA")

p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(style = "equal", n = 10,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Distribution of total water points by LGA")

tmap_arrange(p2, p1, nrow = 1)
```

## Rate Map

### Why Map Rates Instead of Counts?

Mapping raw counts is misleading because larger LGAs naturally have more water points. A rate map shows the proportion of functional water points, which is comparable across LGAs of different sizes.

```{r}
NGA_wp <- NGA_wp |>
  mutate(pct_functional    = wp_functional / total_wp,
         pct_nonfunctional = wp_nonfunctional / total_wp)
```

```{r}
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(style = "equal", n = 10,
                                              values = "brewer.blues"),
              fill.legend = tm_legend(position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1, fill_alpha = 1) +
  tm_title("Rate map of functional water points by LGA")
```

## Extreme Value Maps

Extreme value maps are designed to highlight outliers, specifically the areas at the very top and bottom of the distribution that standard classification methods tend to hide.

## Percentile Map

A percentile map uses 6 fixed categories: 0–1%, 1–10%, 10–50%, 50–90%, 90–99%, 99–100%.

### Helper Functions

```{r}
# Extract a variable from an sf data frame as a plain vector (drops geometry)
get.var <- function(vname, df) {
  v <- df[vname] |> st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

```{r}
# Draw a percentile map for any variable
percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var     <- get.var(vnam, df)
  bperc   <- quantile(var, percent)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_polygons(vnam,
                title   = legtitle,
                breaks  = bperc,
                palette = "Blues",
                labels  = c("< 1%", "1%–10%", "10%–50%",
                            "50%–90%", "90%–99%", "> 99%")) +
    tm_borders() +
    tm_layout(main.title      = mtitle,
              title.position = c("right", "bottom"))
}
```

```{r}
# Remove NAs first
NGA_wp <- NGA_wp |> drop_na()

# Draw the percentile map
percentmap("total_wp", NGA_wp)
```

## Box Map

A box map is an augmented quartile map that explicitly flags **statistical outliers** using IQR fences — equivalent to a boxplot drawn on a map.

### Helper Functions

```{r}
# Compute 7 break points for a box map including outlier fences
boxbreaks <- function(v, mult = 1.5) {
  qv     <- unname(quantile(v))
  iqr    <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  bb <- vector(mode = "numeric", length = 7)
  if (lofence < qv[1]) { bb[1] <- lofence; bb[2] <- floor(qv[1])   }
  else                  { bb[2] <- lofence; bb[1] <- qv[1]           }
  if (upfence > qv[5]) { bb[7] <- upfence; bb[6] <- ceiling(qv[5]) }
  else                  { bb[6] <- upfence; bb[7] <- qv[5]           }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

```{r}
# Draw a box map for any variable
boxmap <- function(vnam, df, legtitle = NA, mtitle = "Box Map", mult = 1.5) {
  var <- get.var(vnam, df)
  bb  <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(vnam,
            title   = legtitle,
            breaks  = bb,
            palette = "Blues",
            labels  = c("Lower outlier", "< 25%", "25%–50%",
                        "50%–75%", "> 75%", "Upper outlier")) +
    tm_borders() +
    tm_layout(main.title      = mtitle,
              title.position = c("left", "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

Shaded areas labelled "Upper outlier" are LGAs with disproportionately high numbers of non-functional water points, which are exactly the locations that warrant further investigation.
