---
title: "Hands-on Exercise 2: Beyond ggplot2 Fundamentals"
author: "carina peh"
date: "April 22, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
format: html
---

## Getting Started

### Loading the packages

```{r}
pacman::p_load(ggrepel, patchwork, 
               ggthemes, hrbrthemes,
               tidyverse)
```

### Importing data

```{r}
exam_data <- read_csv("Exam_data.csv")
```

## Beyond ggplot2 Annotation: ggrepel

### Without ggrepel (messy labels)

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### With ggrepel (clean labels)

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## Beyond ggplot2 Themes

### Default theme_gray()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  theme_gray() +
  ggtitle("Distribution of Maths scores")
```

### Using ggthemes: theme_economist()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_economist()
```

### Using hrbrthemes: theme_ipsum()

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum()
```

### theme_ipsum() with custom settings

```{r}
ggplot(data=exam_data, 
       aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## Beyond Single Graph

### Creating the individual plots

```{r}
p1 <- ggplot(data=exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") + 
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of Maths scores")

p2 <- ggplot(data=exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins=20, 
                 boundary = 100,
                 color="grey25", 
                 fill="grey90") +
  coord_cartesian(xlim=c(0,100)) +
  ggtitle("Distribution of English scores")

p3 <- ggplot(data=exam_data, 
             aes(x= MATHS, 
                 y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, size=0.5) +  
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

### Combining two graphs

```{r}
p1 + p2
```

### Combining three graphs

```{r}
(p1 / p2) | p3
```

### With tags

```{r}
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'I')
```

### With inset

```{r}
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

### patchwork + ggtheme combined

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## Extra Alternatives

### Alternative 1: Using theme_wsj()

I used theme_wsj() instead of theme_economist() for a cleaner, more minimal Wall Street Journal style.

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=20, color="grey25", fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_wsj()
```

### Alternative 2: Using theme_fivethirtyeight()

This theme mimics the style used by FiveThirtyEight, great for clean data journalism visuals.

```{r}
ggplot(data=exam_data, aes(x = MATHS)) +
  geom_histogram(bins=20, color="grey25", fill="grey90") +
  ggtitle("Distribution of Maths scores") +
  theme_fivethirtyeight()
```

### Alternative 3: Different patchwork layout

Instead of stacking p1/p2 on the left, I place p1 on top spanning full width for better visual hierarchy.

```{r}
p1 / (p2 | p3)
```

### Alternative 4: Letter tags instead of Roman numerals

Using letter tags (A, B, C) makes subplots easier to reference in written explanations.

```{r}
((p1 / p2) | p3) + 
  plot_annotation(tag_levels = 'A')
```

### Alternative 5: Inset in different position

Placing the inset in the bottom right instead of top left avoids covering the main data cluster.

```{r}
p3 + inset_element(p2,
                   left = 0.5,
                   bottom = 0.0,
                   right = 1.0,
                   top = 0.4)
```
