---
title: "7b. Visualising and Analysing Text Data with R"
author: "carina peh"
date: "May 30, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---

## Overview

Text visualisation transforms unstructured text data into visual representations to support understanding, grouping, comparison and correlation of textual content. In this exercise, the **tidytext** framework is used to process and visualise newsgroup messages from the 20 Newsgroups dataset, covering techniques such as word clouds, TF-IDF analysis and bigram network graphs.

## Reflections

The key takeaway from this exercise is that text data requires significant preprocessing before meaningful visualisation is possible — removing headers, signatures and stop words is just as important as the visualisation itself. TF-IDF is particularly useful because it surfaces words that are distinctive to a newsgroup rather than just frequent across all documents. Visualising bigrams as a network graph was an effective way to reveal common two-word phrases and their relationships.

## Getting Started

### Loading packages

```{r}
pacman::p_load(tidytext, widyr, wordcloud, DT, 
               ggwordcloud, textplot, lubridate, hms,
               tidyverse, tidygraph, ggraph, igraph)
```

## Importing Multiple Text Files from Multiple Folders

### Creating a folder list

```{r}
news20 <- "20news/"
```

### Defining a function to read files

A custom function `read_folder()` is written to read all text files 
from a folder into a single data frame. `map()` of the **purrr** 
package is used to apply the function to each file, while `unnest()` 
flattens the resulting list into a data frame.

```{r}
read_folder <- function(infolder) {
  tibble(file = dir(infolder,
                    full.names = TRUE)) %>%
    mutate(text = map(file,
                      read_lines)) %>%
    transmute(id = basename(file),
              text) %>%
    unnest(text)
}
```

### Reading in all messages from the 20news folder

```{r}
raw_text <- tibble(folder =
                     dir(news20,
                         full.names = TRUE)) %>%
  mutate(folder_out = map(folder,
                          read_folder)) %>%
  unnest(cols = c(folder_out)) %>%
  mutate(newsgroup = basename(folder)) %>%
  select(newsgroup, id, text)

write_rds(raw_text, "20news.rds")
```

## Initial EDA

The bar chart below shows the frequency of messages by newsgroup.

::: panel-tabset
## The plot
```{r}
#| echo: false
raw_text %>%
  group_by(newsgroup) %>%
  summarize(messages = n_distinct(id)) %>%
  ggplot(aes(messages, newsgroup)) +
  geom_col(fill = "lightblue") +
  labs(y = NULL)
```
## The code chunk
```{r}
#| eval: false
raw_text %>%
  group_by(newsgroup) %>%
  summarize(messages = n_distinct(id)) %>%
  ggplot(aes(messages, newsgroup)) +
  geom_col(fill = "lightblue") +
  labs(y = NULL)
```
:::

## Introducing tidytext

### Removing headers and automated email signatures

Each message contains a header and automated email signature that are not useful for analysis. `cumsum()` of base R is used to return a cumulative sum, while `str_detect()` of **stringr** detects the presence of a pattern in a string.

```{r}
cleaned_text <- raw_text %>%
  group_by(newsgroup, id) %>%
  filter(cumsum(text == "") > 0,
         cumsum(str_detect(
           text, "^--")) == 0) %>%
  ungroup()
```

### Removing lines with nested quotes

Regular expressions are used to remove lines with nested text representing quotes from other users.

```{r}
cleaned_text <- cleaned_text %>%
  filter(str_detect(text, "^[^>]+[A-Za-z\\d]")
         | text == "",
         !str_detect(text,
                     "writes(:|\\.\\.\\.)$"),
         !str_detect(text,
                     "^In article <")
  )
```

### Text data processing

`unnest_tokens()` of **tidytext** splits each message into individual words (tokens). Stop words — common words like "the" and "and" — are then removed using `stop_words`.

```{r}
usenet_words <- cleaned_text %>%
  unnest_tokens(word, text) %>%
  filter(str_detect(word, "[a-z']$"),
         !word %in% stop_words$word)
```

Counting individual words across the entire dataset:

```{r}
usenet_words %>%
  count(word, sort = TRUE)
```

Counting words within each newsgroup:

```{r}
words_by_newsgroup <- usenet_words %>%
  count(newsgroup, word, sort = TRUE) %>%
  ungroup()
```

### Visualising words in newsgroups — wordcloud

`wordcloud()` of the **wordcloud** package is used to plot a static 
wordcloud.

::: panel-tabset
## The plot
```{r}
#| echo: false
#| warning: false
wordcloud(words_by_newsgroup$word,
          words_by_newsgroup$n,
          max.words = 300)
```
## The code chunk
```{r}
#| eval: false
wordcloud(words_by_newsgroup$word,
          words_by_newsgroup$n,
          max.words = 300)
```
:::

### Visualising words in newsgroups — ggwordcloud

`geom_text_wordcloud()` of **ggwordcloud** is used to plot a wordcloud 
faceted by newsgroup.

::: panel-tabset
## The plot
```{r}
#| echo: false
#| warning: false
set.seed(1234)

words_by_newsgroup %>%
  filter(n > 5) %>%
  ggplot(aes(label = word,
             size = n)) +
  geom_text_wordcloud() +
  theme_minimal() +
  facet_wrap(~newsgroup)
```

## The code chunk
```{r}
#| echo: false
#| warning: false
set.seed(1234)

words_by_newsgroup %>%
  filter(n > 5) %>%
  ggplot(aes(label = word,
             size = n)) +
  geom_text_wordcloud() +
  theme_minimal() +
  facet_wrap(~newsgroup)
```
:::

## TF-IDF Analysis

### Computing TF-IDF within newsgroups

`bind_tf_idf()` of **tidytext** computes and binds the term frequency, 
inverse document frequency and TF-IDF of a tidy text dataset.

```{r}
tf_idf <- words_by_newsgroup %>%
  bind_tf_idf(word, newsgroup, n) %>%
  arrange(desc(tf_idf))
```

### Visualising TF-IDF as interactive table

`datatable()` of **DT** creates an interactive table with pagination 
and filtering. `formatRound()` customises the decimal places and 
`formatStyle()` adjusts the row height.

::: panel-tabset
## The plot
```{r}
#| echo: false
DT::datatable(tf_idf, filter = 'top') %>%
  formatRound(columns = c('tf', 'idf',
                          'tf_idf'),
              digits = 3) %>%
  formatStyle(0,
              target = 'row',
              lineHeight='25%')
```
## The code chunk
```{r}
#| eval: false
DT::datatable(tf_idf, filter = 'top') %>%
  formatRound(columns = c('tf', 'idf',
                          'tf_idf'),
              digits = 3) %>%
  formatStyle(0,
              target = 'row',
              lineHeight='25%')
```
:::

### Visualising TF-IDF within newsgroups

::: panel-tabset
## The plot
```{r}
#| echo: false
tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf, n = 12) %>%
  ungroup() %>%
  mutate(word = reorder(word, tf_idf)) %>%
  ggplot(aes(tf_idf, word, fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ newsgroup, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```
## The code chunk
```{r}
#| eval: false
tf_idf %>%
  filter(str_detect(newsgroup, "^sci\\.")) %>%
  group_by(newsgroup) %>%
  slice_max(tf_idf, n = 12) %>%
  ungroup() %>%
  mutate(word = reorder(word, tf_idf)) %>%
  ggplot(aes(tf_idf, word, fill = newsgroup)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ newsgroup, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```
:::

## Word Correlation Analysis

### Counting and correlating pairs of words

`pairwise_cor()` of **widyr** is used to find words that are 
correlated with each other within the same newsgroup.

```{r}
newsgroup_cors <- words_by_newsgroup %>%
  pairwise_cor(newsgroup,
               word,
               n,
               sort = TRUE)
```

### Visualising correlation as a network

::: panel-tabset
## The plot
```{r}
#| echo: false
set.seed(2017)

newsgroup_cors %>%
  filter(correlation > .025) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation,
                     width = correlation)) +
  geom_node_point(size = 6,
                  color = "lightblue") +
  geom_node_text(aes(label = name),
                 repel = TRUE) +
  theme_graph()
```
## The code chunk
```{r}
#| eval: false
set.seed(2017)

newsgroup_cors %>%
  filter(correlation > .025) %>%
  graph_from_data_frame() %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(alpha = correlation,
                     width = correlation)) +
  geom_node_point(size = 6,
                  color = "lightblue") +
  geom_node_text(aes(label = name),
                 repel = TRUE) +
  theme_graph()
```
:::

## Bigram Analysis

### Counting bigrams

A bigram is a pair of consecutive words. `unnest_tokens()` with 
`token = "ngrams"` and `n = 2` extracts all bigrams from the text.

```{r}
bigrams <- cleaned_text %>%
  unnest_tokens(bigram,
                text,
                token = "ngrams",
                n = 2)

bigrams_count <- bigrams %>%
  count(bigram, sort = TRUE)
```

### Cleaning bigrams

Bigrams containing stop words are removed by separating each bigram 
into two words and filtering out those containing stop words.

```{r}
bigrams_separated <- bigrams_count %>%
  separate(bigram, c("word1", "word2"),
           sep = " ")

bigrams_filtered <- bigrams_separated %>%
  filter(!word1 %in% stop_words$word) %>%
  filter(!word2 %in% stop_words$word)
```

### Counting bigrams again

`unite()` of **tidyr** is used to combine the two word columns back 
into a single bigram column.

```{r}
bigrams_united <- bigrams_filtered %>%
  unite(bigram, word1, word2, sep = " ")

bigrams_united
```

### Creating a network graph from bigram data

`graph_from_data_frame()` of **igraph** converts the bigram data into 
a network object.

```{r}
bigram_graph <- bigrams_filtered %>%
  filter(n > 3) %>%
  filter(!is.na(word1), !is.na(word2)) %>%
  graph_from_data_frame()

bigram_graph
```

### Visualising bigrams with ggraph

::: panel-tabset
## The plot
```{r}
#| echo: false
set.seed(1234)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point() +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1)
```
## The code chunk
```{r}
#| eval: false
set.seed(1234)

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link() +
  geom_node_point() +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1)
```
:::

### Revised bigram network graph

::: panel-tabset
## The plot
```{r}
#| echo: false
set.seed(1234)

a <- grid::arrow(type = "closed",
                 length = unit(.15, "inches"))

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n),
                 show.legend = FALSE,
                 arrow = a,
                 end_cap = circle(.07, 'inches')) +
  geom_node_point(color = "lightblue",
                  size = 5) +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1) +
  theme_void()
```
## The code chunk
```{r}
#| eval: false
set.seed(1234)

a <- grid::arrow(type = "closed",
                 length = unit(.15, "inches"))

ggraph(bigram_graph, layout = "fr") +
  geom_edge_link(aes(edge_alpha = n),
                 show.legend = FALSE,
                 arrow = a,
                 end_cap = circle(.07, 'inches')) +
  geom_node_point(color = "lightblue",
                  size = 5) +
  geom_node_text(aes(label = name),
                 vjust = 1,
                 hjust = 1) +
  theme_void()
```
:::