---
title: "9a. Choropleth Mapping with R"
description: "Choropleth mapping visualises spatial distributions using graduated colours applied to enumeration units such as planning subzones, using the tmap and sf packages in R."
author: "carina peh"
date: "June 20, 2026"
date-modified: "last-modified"
execute:
  eval: false
  echo: true
  warning: false
  freeze: true
---

## Overview

Choropleth mapping involves the symbolisation of enumeration units (such as countries, provinces, states, counties or census units) using area patterns or graduated colours. In this exercise, the `tmap` package is used to plot functional and truthful choropleth maps showing the spatial distribution of the aged population across Singapore's planning subzones.

## Reflections

This exercise revealed how much design thinking goes into a choropleth map. The choice of data classification method (e.g. quantile vs equal vs jenks) changes the story the map tells entirely. Quantile classification distributes observations evenly across classes, while equal intervals can make sparse distributions look misleading. Colour palette direction also matters: reversing a palette (using `-brewer.greens`) can flip the visual emphasis completely. The key takeaway is that choropleth maps are powerful but easy to mislead with. Every design choice carries analytical responsibility.

## Getting Started

### Loading Packages

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

**What each package does:**

- `sf` handles geospatial data using the Simple Features standard
- `tmap` creates thematic maps including choropleth and proportional symbol maps
- `tidyverse` covers data wrangling with dplyr, tidyr and readr

## Importing Data into R

### Importing Geospatial Data

```{r}
mpsz <- st_read(dsn = "MP14_SUBZONE_WEB_PL",
                layer = "MP14_SUBZONE_WEB_PL")
```

- `st_read()` from the **sf** package imports the ESRI shapefile as a simple feature data frame
- The dataset contains 323 planning subzones with their boundary geometries in SVY21 coordinate reference system

### Importing Attribute Data

```{r}
popdata <- read_csv("respopagesextod2011to2020.csv")
```

### Data Preparation

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(POP = sum(Pop)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG, values_from = POP) %>%
  mutate(YOUNG            = rowSums(.[3:6]) + rowSums(.[12]),
         `ECONOMY ACTIVE` = rowSums(.[7:11]) + rowSums(.[13:15]),
         AGED             = rowSums(.[16:21]),
         TOTAL            = rowSums(.[3:21]),
         DEPENDENCY       = (YOUNG + AGED) / `ECONOMY ACTIVE`) %>%
  select(PA, SZ, YOUNG, `ECONOMY ACTIVE`, AGED, TOTAL, DEPENDENCY)
```

### Joining Geospatial and Attribute Data

```{r}
# Standardise case before joining
popdata2020 <- popdata2020 |>
  mutate(across(c(PA, SZ), toupper)) |>
  filter(`ECONOMY ACTIVE` > 0)

# Join using mpsz as the left table so the output remains a simple feature
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                           by = c("SUBZONE_N" = "SZ"))

# Save for reuse
write_rds(mpsz_pop2020, "mpszpop2020.rds")
```

## Choropleth Mapping with tmap

### Quick Map with qtm()

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, fill = "DEPENDENCY")
```

- `tmap_mode("plot")` produces a static map. Use `"view"` for an interactive mode instead
- `qtm()` is the quickest way to draw a choropleth map, useful for initial exploration

### Detailed Map with tmap Elements

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style  = "quantile",
                n      = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by Planning Subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: URA and DOS",
             position = c("left", "bottom"))
```

**What each element does:**

- `tm_shape()` defines the spatial data to use
- `tm_polygons()` fills and borders the polygons
- `tm_scale_intervals()` sets the classification method and colour palette
- `tm_compass()`, `tm_grid()` and `tm_credits()` add cartographic furniture like compass, grid and credits

## Data Classification Methods

```{r}
# Jenks natural breaks
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

```{r}
# Equal interval
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

**Key insight:** Quantile classification distributes observations evenly across classes; equal interval classification can mask sparse distributions.

## Small Multiple Choropleth Maps

```{r}
# Side-by-side comparison using tmap_arrange()
youngmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("YOUNG", style = "quantile", palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("AGED", style = "quantile", palette = "Blues")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

Small multiple maps allow visual comparison of how two spatial distributions differ across the same geographic area.
