---
title: "Take-home Exercise 1: Uncovering Loss Drivers in Motor Vehicle Insurance Portfolio using Data Visualisation"
author: "Carina Peh"
date: "May 17, 2026"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
  code-fold: true
---

# 1 Overview

## 1.1 Background

The Property & Casualty (P&C) insurance industry has long struggled withunderperforming portfolios where the root causes of poor profitability and high volatility remain unknown. Traditional approaches to data-driven decision-making in insurance have largely focused on dashboard reporting, which is limited by the general lack of advanced analytical capabilities which offers a more powerful alternative. 
By combining data science,  risk knowledge and industry expertise, insurers can generate actionable insights to grow, optimise existing portfolios and improve operational efficiency. As noted bythe Swiss Re Institute (2019), leveraging visual analytics can potentially help  investigate trends in underlying loss drivers, while data enrichment using external  data can help refine segmentation and underwriting strategy.

In this study, a motor vehicle insurance portfolio dataset from a Spanish insurance company, made publicly available via Mendeley Data, is used to demonstrate the potential of visual analytics techniques in exploring and analysing key factors influencing portfolio profitability and volatility.

## 1.2 The Data

The data for this exercise is the motor vehicle insurance portfolio dataset retrieved from [Mendeley Data](https://data.mendeley.com/datasets/sw4jmdb2sm/1).


The dataset contains **354,140 policy records** across **47 variables**, covering the period from 2022 to 2024. Variables are grouped into four broad categories:

| Category | Description |
|---|---|
| Policy & Insured characteristics | Policy type, status, business origin, payment frequency, bonus score |
| Driver & Vehicle characteristics | Driver age, vehicle age, fuel type, vehicle brand, vehicle value, area |
| Premiums | Total and coverage-level premiums collected |
| Exposure, Claims & Incurred | Claims counts, incurred costs, and exposure duration |

## 1.3 The Task

Using appropriate **Exploratory Data Analysis (EDA)** methods and **ggplot2** functions, this study aims to:

- Apply suitable **univariate** visual analytics techniques to explore the distributions and characteristics of numerical and categorical variables.

- Apply suitable **bivariate** visual analytics techniques to investigate and statistically validate factors influencing the **profitability** and **volatility** of the insurance portfolio.

# 2 Loading Packages

| Package | Description |
|---|---|
| [**tidyverse**](https://www.tidyverse.org/) | A collection of R packages designed for data science, used for data import, wrangling and transformation |
| [**readxl**](https://readxl.tidyverse.org/) | For reading Excel files such as the variable cookbook |
| [**knitr**](https://yihui.org/knitr/) | For dynamic report generation and displaying formatted tables |
| [**patchwork**](https://patchwork.data-imaginist.com/) | For combining multiple ggplot2 plots into a single composite figure |
| [**ggthemes**](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/) | Extra themes, geoms and scales for ggplot2 |
| [**scales**](https://scales.r-lib.org/) | For formatting axis labels and annotations in ggplot2 |
| [**ggridges**](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) | For creating ridgeline plots to visualise distributions |
| [**ggdist**](https://mjskay.github.io/ggdist/) | For visualising distributions and uncertainty |
| [**ggpubr**](https://rpkgs.datanovia.com/ggpubr/) | For creating publication-ready ggplot2 plots |
| [**ggstatsplot**](https://indrajeetpatil.github.io/ggstatsplot/) | For combining statistical tests with visualisations |

The code chunk below uses `p_load()` of the [**pacman**](https://cran.r-project.org/web/packages/pacman/vignettes/Introduction_to_pacman.html) 
package to check if the packages are installed. If they are, they will be loaded into R. If not, they will be installed first before being loaded.

```{r}
pacman::p_load(tidyverse, readxl, knitr,
               patchwork, ggthemes, scales,
               ggridges, ggdist, ggpubr,
               ggstatsplot, naniar)
```

# 3 Importing the Data

In this section, the following steps are taken to import and perform initial checks on the data, from reading in the raw data to verifying its structure. The **tidyverse** package will be used, specifically **readr** for reading in the data and **dplyr** for data manipulation.

The code chunk below imports the motor vehicle insurance portfolio dataset using `read_delim()` of the **readr** package. The dataset uses a semicolon (`;`) as its delimiter instead of a comma, so `delim = ";"` is specified.

```{r}
insurance <- read_delim("Dataset of motor insurance portfolio.csv",
                        delim = ";",
                        show_col_types = FALSE)
```

The dataset is in a tibble dataframe, containing **354,140 observations (rows)** across **47 variables (columns)**. Each observation corresponds to a motor vehicle insurance policy record from a Spanish non-life insurance company.

We use `glimpse()` of the **dplyr** package to get a quick overview of the dataset structure, including the number of rows, columns and data types of each variable.

```{r}
glimpse(insurance)
```

Before moving on, duplicated records are checked to prevent duplicates in our analysis using `group_by_all()`.

```{r}
duplicate <- insurance %>%
  group_by_all() %>%
  filter(n() > 1) %>%
  ungroup()

duplicate
```

Missing values are also checked across all columns using `colSums()` and `is.na()`.

```{r}
colSums(is.na(insurance))
```

The results show that four variables contain missing values — `fuel_type` (1,287), `vehicle_value` (513), `vehicle_age` (2) and `age_driving_licence` (2). These will be handled in the Data Wrangling section.

# 4 Data Wrangling

The following flowchart shows the main variables used and created in this study.

```{mermaid}
flowchart LR
    A(Motor Insurance\nPortfolio) -->|Select| B[Policy &\nInsured\nCharacteristics]
    A -->|Select| C[Driver & Vehicle\nCharacteristics]
    A -->|Select| D[Premiums]
    A -->|Select| E[Exposure,\nClaims &\nIncurred]

    B --> B1([policy_type])
    B --> B2([policy_status])
    B --> B3([business_type])
    B --> B4([payment_frequency])
    B --> B5([bonus_score])

    C --> C1([driver_age])
    C --> C2([vehicle_age])
    C --> C3([vehicle_value])
    C --> C4([fuel_type])
    C --> C5([vehicle_brand])
    C --> C6([municipality_type])
    C --> C7([circulation_area])

    D --> D1([total_premium])
    D --> D2([liability_premium])
    D --> D3([property_damage_premium])
    D --> D4([theft_premium])
    D --> D5([fire_premium])
    D --> D6([glass_premium])
    D --> D7([legal_protection_premium])
    D --> D8([occupants_premium])

    E --> E1([total_exposure])
    E --> E2([total_claims])
    E --> E3([total_incurred])
    E --> E4([loss_ratio])
    E --> E5([claim_frequency])
    E --> E6([has_claim])
```

## 4.1 Variable Selection

Below are the variables retained for this study, which can be grouped into 4 broad categories of Policy & Insured Characteristics, Driver & Vehicle  Characteristics, Premiums, and Exposure, Claims & Incurred. Details are extracted from the cookbook(`Descriptive_of_variables.xlsx`):

| Category | Variable | Description | Type | Encoding |
|---|---|---|---|---|
| Policy & Insured | `policy_type` | Main coverage structure | CAT | TP, TPG, CC, COMP_E, COMP_N |
| Policy & Insured | `policy_status` | Policy active or cancelled | CAT | A - Active, C - Cancelled |
| Policy & Insured | `business_type` | Business origin | CAT | NB - New Business, P - Portfolio |
| Policy & Insured | `payment_frequency` | Billing frequency | CAT | A - Annual, S - Semi-annual, Q - Quarterly |
| Policy & Insured | `bonus_score` | Past claims experience | CAT | G - Good, N - Neutral, B - Bad |
| Driver & Vehicle | `driver_age` | Age of main driver | NUM | In years |
| Driver & Vehicle | `vehicle_age` | Age of insured vehicle | NUM | In years |
| Driver & Vehicle | `vehicle_value` | Declared insured value | NUM | In euros (€) |
| Driver & Vehicle | `fuel_type` | Vehicle fuel type | CAT | D - Diesel, G - Gasoline |
| Driver & Vehicle | `vehicle_brand` | Brand of vehicle | CAT | 65 brands |
| Driver & Vehicle | `municipality_type` | Geographic classification | CAT | I - Inland, C - Coastal, IS - Islands |
| Driver & Vehicle | `circulation_area` | Circulation environment | CAT | U - Urban, R - Rural |
| Premiums | `total_premium` | Total premium collected | NUM | In euros (€) |
| Premiums | `liability_premium` | Liability coverage premium | NUM | In euros (€) |
| Premiums | `property_damage_premium` | Property damage premium | NUM | In euros (€) |
| Premiums | `theft_premium` | Theft coverage premium | NUM | In euros (€) |
| Premiums | `fire_premium` | Fire coverage premium | NUM | In euros (€) |
| Premiums | `glass_premium` | Glass coverage premium | NUM | In euros (€) |
| Premiums | `legal_protection_premium` | Legal protection premium | NUM | In euros (€) |
| Premiums | `occupants_premium` | Occupants coverage premium | NUM | In euros (€) |
| Exposure, Claims & Incurred | `total_exposure` | Time-on-risk in policy years | NUM | 0 to 1 |
| Exposure, Claims & Incurred | `total_claims` | Total claims reported | NUM | Count |
| Exposure, Claims & Incurred | `total_incurred` | Total incurred cost of claims | NUM | In euros (€) |
| Exposure, Claims & Incurred | `loss_ratio`* | Profitability measure | NUM | total_incurred / total_premium |
| Exposure, Claims & Incurred | `claim_frequency`* | Volatility measure | NUM | total_claims / total_exposure |
| Exposure, Claims & Incurred | `has_claim`* | Binary claim indicator | NUM | 1 if claims > 0, else 0 |

*Derived variables created in Section 4.5.

The code chunk below uses `select()` of the **dplyr** package to retain only the variables relevant to this study.

```{r}
insurance <- insurance %>%
  select(insured_id, year,
         
         # Policy characteristics
         policy_type, policy_status, business_type,
         payment_frequency, bonus_score,
         
         # Driver & vehicle characteristics
         driver_age, vehicle_age, age_driving_licence,
         fuel_type, vehicle_value, seats,
         power_to_weight_ratio, vehicle_brand,
         municipality_type, circulation_area,
         
         # Premiums
         total_premium, liability_premium,
         property_damage_premium, theft_premium,
         fire_premium, glass_premium,
         legal_protection_premium, occupants_premium,
         
         # Exposure, claims & incurred
         total_exposure, liability_exposure,
         total_claims, total_incurred)
```

The resulting dataset retains **354,140 observations** across **30 variables**.


## 4.2 Handling Missing Values

`is.na()` allows us to identify missing values in our dataset and `sum()` counts the frequency of such missing values.

```{r}
# Checks for NA in each column
sum(is.na(insurance$fuel_type))
sum(is.na(insurance$vehicle_value))
sum(is.na(insurance$vehicle_age))
sum(is.na(insurance$age_driving_licence))
```

`gg_miss_var()` of the **naniar** package allows us to visualise the missing values across all variables at a glance.

```{r}
gg_miss_var(insurance)
```

The results show that four variables contain missing values:

| Variable | Missing Count | % of Total |
|---|---|---|
| `fuel_type` | 1,287 | 0.36% |
| `vehicle_value` | 513 | 0.14% |
| `vehicle_age` | 2 | <0.01% |
| `age_driving_licence` | 2 | <0.01% |

As the missing values are very small in proportion (less than 0.5% of the total dataset), they will not significantly affect the analysis. `drop_na()` drops rows where any column specified contains a missing value.

```{r}
insurance <- drop_na(insurance,
                     # Drops rows where either of these columns have NAs
                     fuel_type, vehicle_value,
                     vehicle_age, age_driving_licence)

# Here's what the dataset looks like 
insurance
```

The cleaned dataset retains **352,338 observations** — a reduction of only 1,802 rows (0.51%) from the original 354,140, confirming that the removal of missing values has minimal impact on the dataset. Currently, the following code lists the ccolumn names in the dataset.

```{r}
colnames(insurance)
```

## 4.3 Recoding Categorical Variables

`recode()` is primarily used to replace specific values in a vector with new ones. The recoded variable is then assigned using `mutate()` of the 
**dplyr** package.

::: panel-tabset
## Policy Type

`policy_type` is recoded to reflect the full coverage description:

```{r}
insurance <- insurance %>%
  mutate(policy_type = recode(policy_type,
    "TP"     = "Third-Party",
    "TPG"    = "Third-Party with Glass",
    "CC"     = "Third-Party Combined",
    "COMP_E" = "Comprehensive with Excess",
    "COMP_N" = "Comprehensive without Excess"))
```

## Policy Status

`policy_status` is recoded to reflect whether the policy is active 
or cancelled:

```{r}
insurance <- insurance %>%
  mutate(policy_status = recode(policy_status,
    "A" = "Active",
    "C" = "Cancelled"))
```

## Business Type

`business_type` is recoded to reflect the business origin:

```{r}
insurance <- insurance %>%
  mutate(business_type = recode(business_type,
    "NB" = "New Business",
    "P"  = "Portfolio"))
```

## Payment Frequency

`payment_frequency` is recoded to reflect the billing frequency:

```{r}
insurance <- insurance %>%
  mutate(payment_frequency = recode(payment_frequency,
    "A" = "Annual",
    "S" = "Semi-Annual",
    "Q" = "Quarterly"))
```

## Bonus Score

`bonus_score` is recoded to reflect the claims history:

```{r}
insurance <- insurance %>%
  mutate(bonus_score = recode(bonus_score,
    "G" = "Good",
    "N" = "Neutral",
    "B" = "Bad"))
```

## Fuel Type

`fuel_type` is recoded to reflect the vehicle fuel type:

```{r}
insurance <- insurance %>%
  mutate(fuel_type = recode(fuel_type,
    "D" = "Diesel",
    "G" = "Gasoline"))
```

## Municipality Type

`municipality_type` is recoded to reflect the geographic classification:

```{r}
insurance <- insurance %>%
  mutate(municipality_type = recode(municipality_type,
    "I"  = "Inland",
    "C"  = "Coastal",
    "IS" = "Islands"))
```

## Circulation Area

`circulation_area` is recoded to reflect the circulation environment:

```{r}
insurance <- insurance %>%
  mutate(circulation_area = recode(circulation_area,
    "U" = "Urban",
    "R" = "Rural"))
```

:::


## 4.4 Data Type Conversion

From the `glimpse()` output earlier, the categorical variables are currently stored as character (`chr`) data type. They need to be converted to **factors** (`fct`) so that R treats them correctly in visualisations and statistical analyses. `mutate()` and `as.factor()` of the **dplyr** package are used to perform the conversion.

```{r}
insurance <- insurance %>%
  mutate(policy_type       = as.factor(policy_type),
         policy_status     = as.factor(policy_status),
         business_type     = as.factor(business_type),
         payment_frequency = as.factor(payment_frequency),
         bonus_score       = as.factor(bonus_score),
         fuel_type         = as.factor(fuel_type),
         vehicle_brand     = as.factor(vehicle_brand),
         municipality_type = as.factor(municipality_type),
         circulation_area  = as.factor(circulation_area),
         year              = as.factor(year),
         seats             = as.factor(seats))
```

We use `glimpse()` to verify that the data types have been correctly converted.

```{r}
glimpse(insurance)
```

The categorical variables are now stored as factors (`fct`), confirming that the data type conversion was successful.

## 4.5 Variable Creation

Three new variables are created to support the analysis of portfolio profitability and volatility. `mutate()` of the **dplyr** package is used to create these variables.

### 4.5.1 Loss Ratio

`loss_ratio` is calculated as the ratio of total incurred costs to total premium collected. A loss ratio greater than 1 indicates that the insurer paid out more in claims than it collected in premiums, reflecting poor profitability.

```{r}
insurance <- insurance %>%
  mutate(loss_ratio = total_incurred / total_premium)
```

### 4.5.2 Claim Frequency

`claim_frequency` is calculated as the ratio of total claims to total exposure. It measures how often claims occur per policy year, reflecting the volatility of the portfolio.

```{r}
insurance <- insurance %>%
  mutate(claim_frequency = total_claims / total_exposure)
```

### 4.5.3 Has Claim

`has_claim` is a binary indicator variable that takes the value 1 if a policy has at least one claim, and 0 otherwise. It is useful for identifying which policies have experienced a loss.

```{r}
insurance <- insurance %>%
  mutate(has_claim = ifelse(total_claims > 0, 1, 0))
```

We use `glimpse()` to verify that the three newly created variables 
have been successfully added to the dataset.

```{r}
glimpse(insurance)
```

The dataset now contains **352,338 observations** across **32 variables**, 
with `loss_ratio`, `claim_frequency` and `has_claim` successfully added.

The *.rds* file format is usually smaller than its CSV counterpart and will therefore take up less storage space. The *.rds* file will also preserve data types and classes such as factors and numeric variables, eliminating the need to redefine data types after loading the file. For fast and space efficient data storage, the cleaned dataset is exported as an RDS file using `write_rds()` and can be re-imported using `read_rds()`.

```{r}
write_rds(insurance, "insurance_clean.rds")
```

```{r}
insurance <- read_rds("insurance_clean.rds")
```


# 5 Univariate Analysis

## 5.1 Target Variables

The target variables are `loss_ratio`, `claim_frequency`, `total_incurred` and `has_claim`, which measure portfolio profitability, volatility, claim severity and claim occurrence respectively. As some variables contain extreme outliers, the analysis is restricted to a reasonable range for visualisation purposes.Before diving into the Exploratory Data Analysis, the distributions of key variables are examined individually to better understand their characteristics and identify any notable patterns.

::: panel-tabset
## Loss Ratio

`loss_ratio` measures the ratio of total incurred costs to total premium collected. A loss ratio greater than 1 indicates that the insurer paid out more in claims than it collected in premiums.

The raincloud plot below combines a half-violin plot, boxplot and individual data points to provide a comprehensive view of the distribution of `loss_ratio`.

```{r}
#| warning: false
insurance_lr <- insurance %>%
  filter(loss_ratio < 10,
         loss_ratio > 0)

ggplot(insurance_lr,
       aes(x = 0, y = loss_ratio)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA,
               fill = "#4a7c6f") +
  geom_boxplot(width = 0.2,
               outlier.shape = NA,
               fill = "#7aadab",
               alpha = 0.5) +
  coord_flip() +
  labs(title = "Distribution of Loss Ratio",
       subtitle = "Policies with loss ratio between 0 and 10",
       x = NULL,
       y = "Loss Ratio") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

The raincloud plot reveals that `loss_ratio` is highly right-skewed among policies with claims. Most policies record a loss ratio between 1.0 and 2.5, indicating that incurred costs typically exceed premiums collected. The long right tail suggests that a small number of policies experience exceptionally high losses, which significantly impacts overall portfolio profitability.


The histogram and boxplot below provide a complementary view of the distribution of `loss_ratio`.

```{r}
p1 <- ggplot(insurance_lr,
             aes(x = loss_ratio)) +
  geom_histogram(binwidth = 0.1,
                 fill = "#4a7c6f",
                 colour = "white") +
  geom_vline(aes(xintercept = mean(loss_ratio)),
             colour = "red",
             linetype = "dashed",
             linewidth = 0.8) +
  geom_vline(aes(xintercept = median(loss_ratio)),
             colour = "blue",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = mean(insurance_lr$loss_ratio) + 0.3,
           y = Inf, vjust = 2,
           label = paste("Mean:", round(mean(insurance_lr$loss_ratio), 2)),
           colour = "red", size = 3) +
  annotate("text", x = median(insurance_lr$loss_ratio) + 0.3,
           y = Inf, vjust = 4,
           label = paste("Median:", round(median(insurance_lr$loss_ratio), 2)),
           colour = "blue", size = 3) +
  labs(title = "Distribution of Loss Ratio",
       x = "Loss Ratio",
       y = "Count") +
  theme_minimal()

p2 <- ggplot(insurance_lr,
             aes(x = loss_ratio,
                 y = "")) +
  geom_boxplot(fill = "#7aadab",
               outlier.alpha = 0.3,
               outlier.size = 0.5) +
  labs(x = "Loss Ratio",
       y = NULL) +
  theme_minimal()

p2 / p1 +
  plot_layout(heights = c(1, 3.5))
```

The distribution of `loss_ratio` is highly right-skewed, with the majority 
of policies having a loss ratio of 0 (no claims made). The mean loss ratio 
is higher than the median, indicating the presence of a small number of 
policies with very high loss ratios that pull the mean upward.

## Claim Frequency

`claim_frequency` measures the ratio of total claims to total exposure. 
A higher claim frequency indicates greater volatility in the portfolio.

The raincloud plot below provides a comprehensive view of the distribution 
of `claim_frequency`.

```{r}
#| warning: false
insurance_cf <- insurance %>%
  filter(claim_frequency < 10,
         claim_frequency > 0)

ggplot(insurance_cf,
       aes(x = 0, y = claim_frequency)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA,
               fill = "#4a7c6f") +
  geom_boxplot(width = 0.2,
               outlier.shape = NA,
               fill = "#7aadab",
               alpha = 0.5) +
  stat_dots(side = "left",
            justification = 1.1,
            binwidth = 0.05,
            colour = "#4a7c6f",
            alpha = 0.3) +
  coord_flip() +
  labs(title = "Distribution of Claim Frequency",
       subtitle = "Policies with claim frequency between 0 and 10",
       x = NULL,
       y = "Claim Frequency") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

The histogram and boxplot below provide a complementary view of the distribution of `claim_frequency`.

```{r}
p3 <- ggplot(insurance_cf,
             aes(x = claim_frequency)) +
  geom_histogram(binwidth = 0.1,
                 fill = "#4a7c6f",
                 colour = "white") +
  geom_vline(aes(xintercept = mean(claim_frequency)),
             colour = "red",
             linetype = "dashed",
             linewidth = 0.8) +
  geom_vline(aes(xintercept = median(claim_frequency)),
             colour = "blue",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = mean(insurance_cf$claim_frequency) + 0.3,
           y = Inf, vjust = 2,
           label = paste("Mean:", round(mean(insurance_cf$claim_frequency), 2)),
           colour = "red", size = 3) +
  annotate("text", x = median(insurance_cf$claim_frequency) + 0.3,
           y = Inf, vjust = 4,
           label = paste("Median:", round(median(insurance_cf$claim_frequency), 2)),
           colour = "blue", size = 3) +
  labs(title = "Distribution of Claim Frequency",
       x = "Claim Frequency",
       y = "Count") +
  theme_minimal()

p4 <- ggplot(insurance_cf,
             aes(x = claim_frequency,
                 y = "")) +
  geom_boxplot(fill = "#7aadab",
               outlier.alpha = 0.3,
               outlier.size = 0.5) +
  labs(x = "Claim Frequency",
       y = NULL) +
  theme_minimal()

p4 / p3 +
  plot_layout(heights = c(1, 3.5))
```

The distribution of `claim_frequency` is heavily right-skewed, with most policies recording a claim frequency of around 1, meaning one claim per policy year. The mean of 1.95 is noticeably higher than the median of 1.28, driven by a small number of high-frequency claimants. The boxplot reveals numerous outliers beyond the upper whisker, indicating that a subset of policies experience a disproportionately high number of claims, contributing significantly to portfolio volatility.

:::


## Total Incurred

`total_incurred` represents the total incurred cost of claims for each policy, including paid amounts and outstanding reserves. Only policies with claims are included in this analysis.

The raincloud plot below provides a comprehensive view of the distribution of `total_incurred`.

```{r}
#| warning: false
insurance_ti <- insurance %>%
  filter(total_incurred > 0,
         total_incurred < 10000)

ggplot(insurance_ti,
       aes(x = 0, y = total_incurred)) +
  stat_halfeye(adjust = 0.5,
               justification = -0.2,
               .width = 0,
               point_colour = NA,
               fill = "#4a7c6f") +
  geom_boxplot(width = 0.2,
               outlier.shape = NA,
               fill = "#7aadab",
               alpha = 0.5) +
  stat_dots(side = "left",
            justification = 1.1,
            binwidth = 50,
            colour = "#4a7c6f",
            alpha = 0.3) +
  coord_flip() +
  labs(title = "Distribution of Total Incurred",
       subtitle = "Policies with total incurred between €0 and €10,000",
       x = NULL,
       y = "Total Incurred (€)") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

The histogram and boxplot below provide a complementary view of the 
distribution of `total_incurred`.

```{r}
p5 <- ggplot(insurance_ti,
             aes(x = total_incurred)) +
  geom_histogram(binwidth = 200,
                 fill = "#4a7c6f",
                 colour = "white") +
  geom_vline(aes(xintercept = mean(total_incurred)),
             colour = "red",
             linetype = "dashed",
             linewidth = 0.8) +
  geom_vline(aes(xintercept = median(total_incurred)),
             colour = "blue",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = mean(insurance_ti$total_incurred) + 500,
           y = Inf, vjust = 2,
           label = paste("Mean: €", round(mean(insurance_ti$total_incurred), 2)),
           colour = "red", size = 3) +
  annotate("text", x = median(insurance_ti$total_incurred) + 500,
           y = Inf, vjust = 4,
           label = paste("Median: €", round(median(insurance_ti$total_incurred), 2)),
           colour = "blue", size = 3) +
  labs(title = "Distribution of Total Incurred",
       x = "Total Incurred (€)",
       y = "Count") +
  theme_minimal()

p6 <- ggplot(insurance_ti,
             aes(x = total_incurred,
                 y = "")) +
  geom_boxplot(fill = "#7aadab",
               outlier.alpha = 0.3,
               outlier.size = 0.5) +
  labs(x = "Total Incurred (€)",
       y = NULL) +
  theme_minimal()

p6 / p5 +
  plot_layout(heights = c(1, 3.5))
```

The distribution of `total_incurred` is moderately right-skewed, with most policies recording claim costs below €2,500. The mean of €1,271.98 is higher than the median of €898.68, indicating that a small number of high-cost claims pull the average upward. Overall, the majority of claims are relatively manageable in size, though the long tail suggests that occasional large claims do occur within the portfolio.

## Has Claim

`has_claim` is a binary variable indicating whether a policy has recorded 
at least one claim (1) or no claims (0). A bar chart is used to visualise 
the proportion of policies with and without claims.

```{r}
insurance %>%
  mutate(has_claim = ifelse(has_claim == 1, 
                            "Has Claim", 
                            "No Claim")) %>%
  count(has_claim) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = has_claim,
             y = n,
             fill = has_claim)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_fill_manual(values = c("Has Claim" = "#4a7c6f",
                                "No Claim" = "#7aadab")) +
  labs(title = "Distribution of Claim Occurrence",
       x = NULL,
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar chart shows that **85.7%** of policies in the portfolio have not recorded any claims, while only **14.3%** have experienced at least one claim. This is typical of motor insurance portfolios where most policyholders do not make a claim in a given year. Despite being a small proportion, the 14.3% of policies with claims are of particular interest as they are the primary drivers of portfolio profitability and volatility.


## 5.2 Driver Characteristics

This section explores the distribution of driver-related variables in the 
portfolio, specifically `driver_age` and `bonus_score`.

::: panel-tabset
## Driver Age

`driver_age` represents the age of the main driver associated with each 
policy, measured in years. The histogram and boxplot below show the 
distribution of driver ages in the portfolio.

```{r}
p7 <- ggplot(insurance,
             aes(x = driver_age)) +
  geom_histogram(binwidth = 1,
                 fill = "#4a7c6f",
                 colour = "white") +
  geom_vline(aes(xintercept = mean(driver_age)),
             colour = "red",
             linetype = "dashed",
             linewidth = 0.8) +
  geom_vline(aes(xintercept = median(driver_age)),
             colour = "blue",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = mean(insurance$driver_age) + 4,
           y = Inf, vjust = 2,
           label = paste("Mean:", round(mean(insurance$driver_age), 1)),
           colour = "red", size = 3) +
  annotate("text", x = median(insurance$driver_age) + 4,
           y = Inf, vjust = 4,
           label = paste("Median:", round(median(insurance$driver_age), 1)),
           colour = "blue", size = 3) +
  labs(title = "Distribution of Driver Age",
       x = "Driver Age (years)",
       y = "Count") +
  theme_minimal()

p8 <- ggplot(insurance,
             aes(x = driver_age,
                 y = "")) +
  geom_boxplot(fill = "#7aadab",
               outlier.alpha = 0.3,
               outlier.size = 0.5) +
  labs(x = "Driver Age (years)",
       y = NULL) +
  theme_minimal()

p8 / p7 +
  plot_layout(heights = c(1, 3.5))
```

The distribution of `driver_age` is approximately bell-shaped and slightly right-skewed, with most drivers aged between 35 and 65 years. The mean age of 49 and median of 48 are very close, confirming a relatively symmetric distribution. The boxplot shows that the middle 50% of drivers fall between approximately 39 and 58 years old, with a small number of older drivers above 75 appearing as outliers.

## Bonus Score

`bonus_score` reflects the policyholder's past claims experience — Good (G), Neutral (N) or Bad (B). A bar chart is used to visualise the proportion of policies in each category.

```{r}
insurance %>%
  count(bonus_score) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(bonus_score, -n),
             y = n,
             fill = bonus_score)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Good"    = "#4a7c6f",
                                "Neutral" = "#7aadab",
                                "Bad"     = "#d4e8e1")) +
  labs(title = "Distribution of Bonus Score",
       x = "Bonus Score",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none")
```
The bar chart shows that **96.5%** of policies belong to policyholders with a Good bonus score, indicating a strong claims history across the portfolio. Only **2.2%** are Neutral and **1.3%** are Bad. This suggests that the portfolio is largely made up of low-risk, experienced policyholders who have rarely or never made a claim.

:::

## 5.3 Vehicle Characteristics

This section explores the distribution of vehicle-related variables in the portfolio, specifically `vehicle_age`, `vehicle_value`, `fuel_type` and `vehicle_brand`.

::: panel-tabset
## Vehicle Age

`vehicle_age` represents the age of the insured vehicle in years. The histogram and boxplot below show the distribution of vehicle ages in the portfolio.

```{r}
p9 <- ggplot(insurance,
             aes(x = vehicle_age)) +
  geom_histogram(binwidth = 1,
                 fill = "#4a7c6f",
                 colour = "white") +
  geom_vline(aes(xintercept = mean(vehicle_age)),
             colour = "red",
             linetype = "dashed",
             linewidth = 0.8) +
  geom_vline(aes(xintercept = median(vehicle_age)),
             colour = "blue",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = mean(insurance$vehicle_age) + 5,
           y = Inf, vjust = 2,
           label = paste("Mean:", round(mean(insurance$vehicle_age), 1)),
           colour = "red", size = 3) +
  annotate("text", x = median(insurance$vehicle_age) + 5,
           y = Inf, vjust = 4,
           label = paste("Median:", round(median(insurance$vehicle_age), 1)),
           colour = "blue", size = 3) +
  labs(title = "Distribution of Vehicle Age",
       x = "Vehicle Age (years)",
       y = "Count") +
  theme_minimal()

p10 <- ggplot(insurance,
              aes(x = vehicle_age,
                  y = "")) +
  geom_boxplot(fill = "#7aadab",
               outlier.alpha = 0.3,
               outlier.size = 0.5) +
  labs(x = "Vehicle Age (years)",
       y = NULL) +
  theme_minimal()

p10 / p9 +
  plot_layout(heights = c(1, 3.5))
```

The distribution of `vehicle_age` is approximately bell-shaped, with most vehicles aged between 15 and 40 years. The mean of 25.9 years and median of 25 years are very close together, confirming a relatively symmetric distribution. This suggests that the portfolio is dominated by older vehicles, which may reflect the nature of the Spanish non-life insurance market. The boxplot reveals a small number of very old vehicles (over 55 years) appearing as outliers.

## Vehicle Value

`vehicle_value` represents the declared insured value of the vehicle in euros. The histogram and boxplot below show the distribution of vehicle values in the portfolio.

```{r}
p11 <- ggplot(insurance,
              aes(x = vehicle_value)) +
  geom_histogram(binwidth = 2000,
                 fill = "#4a7c6f",
                 colour = "white") +
  geom_vline(aes(xintercept = mean(vehicle_value)),
             colour = "red",
             linetype = "dashed",
             linewidth = 0.8) +
  geom_vline(aes(xintercept = median(vehicle_value)),
             colour = "blue",
             linetype = "dashed",
             linewidth = 0.8) +
  annotate("text", x = mean(insurance$vehicle_value) + 20000,
           y = Inf, vjust = 2,
           label = paste("Mean: €", round(mean(insurance$vehicle_value), 0)),
           colour = "red", size = 3) +
  annotate("text", x = median(insurance$vehicle_value) + 20000,
           y = Inf, vjust = 4,
           label = paste("Median: €", round(median(insurance$vehicle_value), 0)),
           colour = "blue", size = 3) +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Distribution of Vehicle Value",
       x = "Vehicle Value (€)",
       y = "Count") +
  theme_minimal()

p12 <- ggplot(insurance,
              aes(x = vehicle_value,
                  y = "")) +
  geom_boxplot(fill = "#7aadab",
               outlier.alpha = 0.3,
               outlier.size = 0.5) +
  scale_x_continuous(labels = scales::comma) +
  labs(x = "Vehicle Value (€)",
       y = NULL) +
  theme_minimal()

p12 / p11 +
  plot_layout(heights = c(1, 3.5))
```

The distribution of `vehicle_value` is highly right-skewed, with the majority of vehicles valued below €50,000. The mean of €26,937 is slightly higher than the median of €24,687, indicating that a small number of high-value vehicles are pulling the average upward. The boxplot confirms the presence of numerous extreme outliers stretching beyond €200,000, likely representing luxury or high-performance vehicles. Despite these outliers, most policyholders in this portfolio own relatively affordable vehicles.

## Fuel Type

`fuel_type` indicates whether the vehicle runs on Diesel or Gasoline. A bar chart is used to visualise the proportion of each fuel type in the portfolio.

```{r}
insurance %>%
  count(fuel_type) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(fuel_type, -n),
             y = n,
             fill = fuel_type)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Diesel"   = "#4a7c6f",
                                "Gasoline" = "#7aadab")) +
  labs(title = "Distribution of Fuel Type",
       x = "Fuel Type",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar chart shows that **64.2%** of policies are associated with Diesel vehicles, while **35.8%** are Gasoline. The predominance of diesel vehicles reflects the historically strong preference for diesel cars in the Spanish market, particularly for older vehicles which make up a large proportion of this portfolio.

## Vehicle Brand

`vehicle_brand` represents the brand of the insured vehicle. As there are 
65 brands in the dataset, only the top 10 most common brands are shown for 
clarity.

```{r}
insurance %>%
  count(vehicle_brand) %>%
  arrange(desc(n)) %>%
  slice_head(n = 10) %>%
  mutate(pct = n / sum(insurance %>% count(vehicle_brand) %>% pull(n)) * 100) %>%
  ggplot(aes(x = reorder(vehicle_brand, n),
             y = n,
             fill = n)) +
  geom_col(fill = "#4a7c6f") +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            hjust = -0.1,
            size = 3) +
  scale_y_continuous(labels = scales::comma) +
  coord_flip() +
  labs(title = "Top 10 Vehicle Brands by Number of Policies",
       x = "Vehicle Brand",
       y = "Number of Policies") +
  theme_minimal()
```

The bar chart shows that Renault (**9.7%**), Citroen (**9.2%**) and Peugeot (**8.4%**) are the three most common vehicle brands in the portfolio, reflecting the strong presence of French automobile brands in the Spanish market. The top 7 brands are all European, with SEAT being the only Spanish brand in the list. Toyota is the only Asian brand to appear in the top 10, ranking last with **4%** of policies.  Overall, the portfolio is heavily dominated by mainstream European brands, with luxury brands such as Audi and Mercedes accounting for a smaller share.

:::

## 5.4 Policy Characteristics

This section explores the distribution of policy-related variables in the portfolio, specifically `policy_type`, `policy_status`, `business_type` and `payment_frequency`.

::: panel-tabset
## Policy Type

`policy_type` describes the main coverage structure of the motor insurance policy. A bar chart is used to visualise the proportion of each policy type in the portfolio.

```{r}
insurance %>%
  count(policy_type) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(policy_type, -n),
             y = n,
             fill = policy_type)) +
  geom_col(fill = "#4a7c6f") +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Distribution of Policy Type",
       x = "Policy Type",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15,
                                   hjust = 1))
```

The bar chart shows that Third-Party Combined policies account for the largest share of the portfolio at **58%**, followed by Comprehensive with Excess at **25.6%**. Third-Party with Glass (**10.2%**) and Comprehensive without Excess (**4.5%**) make up smaller proportions, while basic Third-Party policies are the least common at only **1.6%**. The dominance of Third-Party Combined policies suggests that most policyholders opt for broader coverage beyond basic liability, combining theft, fire and glass protection in a single policy.

## Policy Status

`policy_status` indicates whether the policy was active or cancelled at the time of analysis. A bar chart is used to visualise the proportion of each policy status.

```{r}
insurance %>%
  count(policy_status) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(policy_status, -n),
             y = n,
             fill = policy_status)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Active"    = "#4a7c6f",
                                "Cancelled" = "#7aadab")) +
  labs(title = "Distribution of Policy Status",
       x = "Policy Status",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar chart shows that **89.2%** of policies are Active, while **10.8%** have been Cancelled. The low cancellation rate suggests a relatively stable and loyal customer base within the portfolio, with most policyholders choosing to maintain their coverage.

## Business Type

`business_type` classifies the policy according to its business origin either New Business (NB) or Portfolio (P). A bar chart is used to visualise the proportion of each business type.

```{r}
insurance %>%
  count(business_type) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(business_type, -n),
             y = n,
             fill = business_type)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("New Business" = "#4a7c6f",
                                "Portfolio"    = "#7aadab")) +
  labs(title = "Distribution of Business Type",
       x = "Business Type",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar chart shows that **69.3%** of policies are New Business, while **30.7%** are Portfolio renewals. The higher proportion of new business suggests that the insurer has been actively growing its customer base during the period covered by the dataset, with roughly 7 out of 10 policies being newly acquired.

## Payment Frequency

`payment_frequency` indicates how often the policy is billed — Annually, Semi-Annually or Quarterly. A bar chart is used to visualise the proportion of each payment frequency.

```{r}
insurance %>%
  count(payment_frequency) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(payment_frequency, -n),
             y = n,
             fill = payment_frequency)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Annual"      = "#4a7c6f",
                                "Semi-Annual" = "#7aadab",
                                "Quarterly"   = "#d4e8e1")) +
  labs(title = "Distribution of Payment Frequency",
       x = "Payment Frequency",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none")
```

The bar chart shows that **81.3%** of policies are billed Annually, followed by Semi-Annual (**14.4%**) and Quarterly (**4.3%**). The overwhelming preference for annual billing suggests that most policyholders are comfortable committing to a full year of coverage upfront, which also benefits the insurer through more stable and  predictable cash flows.

:::

## 5.5 Geographic Characteristics

This section explores the distribution of geographic variables in the portfolio, specifically `municipality_type` and `circulation_area`.

::: panel-tabset
## Municipality Type

`municipality_type` classifies the policyholder's municipality as Inland, Coastal or Islands. A bar chart is used to visualise the proportion of each municipality type in the portfolio.

```{r}
insurance %>%
  count(municipality_type) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(municipality_type, -n),
             y = n,
             fill = municipality_type)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Inland"  = "#4a7c6f",
                                "Coastal" = "#7aadab",
                                "Islands" = "#d4e8e1")) +
  labs(title = "Distribution of Municipality Type",
       x = "Municipality Type",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.margin = margin(t = 30))
```

The bar chart shows that **64.6%** of policies are from Inland municipalities, followed by Coastal (**28.6%**) and Islands (**6.8%**). This reflects the geographic distribution of the Spanish population, where most people live in inland cities such as Madrid. The relatively high proportion of Coastal policies also reflects Spain's extensive coastline and densely populated coastal cities such as Barcelona and 
Valencia.

## Circulation Area

`circulation_area` indicates whether the vehicle predominantly circulates in an Urban or Rural environment. A bar chart is used to visualise the proportion of each circulation area in the portfolio.

```{r}
insurance %>%
  count(circulation_area) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(circulation_area, -n),
             y = n,
             fill = circulation_area)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")),
            vjust = -0.5,
            size = 4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Urban" = "#4a7c6f",
                                "Rural" = "#7aadab")) +
  labs(title = "Distribution of Circulation Area",
       x = "Circulation Area",
       y = "Number of Policies") +
  theme_minimal() +
  theme(legend.position = "none",
        plot.margin = margin(t = 30))
```

The bar chart shows that **53.8%** of policies are associated with Urban circulation, while **46.2%** are Rural. The split is relatively even compared to other variables, suggesting the portfolio covers both urban and rural drivers fairly equally. Urban drivers tend to face higher traffic density, which may lead to more frequent but less severe claims, while rural drivers may experience fewer but potentially more severe accidents due to higher speeds.

:::

# 6 Exploratory Data Analysis

In this section, bivariate visual analytics techniques are applied to investigate and statistically validate the factors influencing the profitability and volatility of the motor insurance portfolio. `ggbetweenstats()` of the **ggstatsplot** package is used to combine statistical tests with visualisations, providing both visual and statistical evidence of the relationships between variables.

## 6.1 Factors Affecting Loss Ratio (Profitability)

`loss_ratio` is used as the measure of portfolio profitability. Only policies with a loss ratio between 0 and 10 are included to exclude extreme outliers that may distort the analysis.

::: panel-tabset
## Bonus Score

The violin plot below examines whether `bonus_score` — reflecting a policyholder's past claims history — has a significant effect on `loss_ratio`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_lr,
  x    = bonus_score,
  y    = loss_ratio,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Loss Ratio by Bonus Score",
  xlab = "Bonus Score",
  ylab = "Loss Ratio",
  palette = "Set2"
)
```

The Kruskal-Wallis test confirms a statistically significant difference in loss ratio across bonus score groups (p = 2.07e-19). Interestingly, policyholders with a **Good** bonus score have the highest median loss ratio of **2.02**, followed by **Neutral** (1.60) and **Bad** (1.44). This counterintuitive finding may be attributed to the fact that Good bonus score policyholders dominate the portfolio (n = 35,663) and tend to hold more comprehensive policies, which attract higher claim costs. All pairwise comparisons are statistically significant, confirming that bonus score is a meaningful differentiator of portfolio profitability.

## Policy Type

The violin plot below examines whether `policy_type` has a significant effect on `loss_ratio`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_lr,
  x    = policy_type,
  y    = loss_ratio,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Loss Ratio by Policy Type",
  xlab = "Policy Type",
  ylab = "Loss Ratio",
  palette = "Set2"
)
```

The Kruskal-Wallis test confirms a statistically significant difference in loss ratio across policy types (p = 5.09e-92). The **Third-Party** policy type has the highest median loss ratio of **4.29** (n = 241), which is notably higher than all other policy types. This may be due to the very small sample size of Third-Party policies, where a few large claims can heavily skew the median upward.

Among the remaining policy types, **Third-Party Combined** (2.25) and **Third-Party with Glass** (2.22) have higher median loss ratios compared to **Comprehensive with Excess** (1.74) and **Comprehensive without Excess** (1.78). This suggests that comprehensive policies, despite offering broader coverage, are associated with lower loss ratios possibly because comprehensive policyholders tend to be more careful drivers who pay higher premiums, resulting in better profitability for the insurer. 


## Fuel Type

The violin plot below examines whether `fuel_type` has a significant effect on `loss_ratio`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_lr,
  x    = fuel_type,
  y    = loss_ratio,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Loss Ratio by Fuel Type",
  xlab = "Fuel Type",
  ylab = "Loss Ratio",
  palette = "Set2"
)
```

The plot examines whether diesel or gasoline vehicles generate different loss ratios.

The Mann-Whitney test confirms a statistically significant difference in loss ratio between fuel types (p = 3.48e-26). Gasoline vehicles have a higher median loss ratio of **2.22** (n = 10,902) compared to Diesel vehicles at **1.92** (n = 26,639). Although the effect size is small (r = -0.07), the difference is statistically significant, suggesting that gasoline vehicles tend to generate slightly less profitable policies than diesel vehicles. This could be attributed to differences in driving behaviour or vehicle characteristics between gasoline and diesel vehicle owners.

## Circulation Area

The violin plot below examines whether `circulation_area` has a significant effect on `loss_ratio`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_lr,
  x    = circulation_area,
  y    = loss_ratio,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Loss Ratio by Circulation Area",
  xlab = "Circulation Area",
  ylab = "Loss Ratio",
  palette = "Set2"
)
```

The Mann-Whitney test confirms a statistically significant difference in loss ratio between circulation areas (p = 2.68e-07). Surprisingly, **Rural** drivers have a slightly higher median loss ratio of **2.06** (n = 15,954) compared to **Urban** drivers at **1.95** (n = 21,587). However, the effect size is very small (r = 0.03), suggesting that while the difference is statistically significant, it is not practically meaningful. This indicates that circulation area alone may not be a strong predictor of portfolio profitability.

## Driver Age

The scatterplot below examines whether `driver_age` has a significant relationship with `loss_ratio`.

```{r}
#| warning: false
#| fig-height: 8
ggscatterstats(
  data = insurance_lr,
  x    = driver_age,
  y    = loss_ratio,
  type = "nonparametric",
  messages = FALSE,
  title = "Loss Ratio by Driver Age",
  xlab = "Driver Age (years)",
  ylab = "Loss Ratio",
  marginal = TRUE
)
```
The Spearman correlation test reveals a statistically significant but very weak positive relationship between driver age and loss ratio (ρ = 0.03, p = 1.00e-07). The nearly flat blue trend line confirms 
that while older drivers tend to have a very slightly higher loss ratio, the relationship is negligible in practical terms. The histograms show that most drivers are aged between 30 and 70, while most loss ratios are concentrated below 2.5. Overall, driver age alone is not a strong predictor of portfolio profitability.

:::

## 6.2 Factors Affecting Claim Frequency (Volatility)

`claim_frequency` is used as the measure of portfolio volatility. Only policies with a claim frequency between 0 and 10 are included to exclude extreme outliers.

::: panel-tabset
## Bonus Score

The violin plot below examines whether `bonus_score` has a significant effect on `claim_frequency`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_cf,
  x    = bonus_score,
  y    = claim_frequency,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Claim Frequency by Bonus Score",
  xlab = "Bonus Score",
  ylab = "Claim Frequency",
  palette = "Set2"
)
```

The Kruskal-Wallis test confirms a statistically significant difference in claim frequency across bonus score groups (p = 8.47e-61). As expected, policyholders with a **Bad** bonus score have the highest median claim frequency of **1.98** (n = 999), followed by **Neutral** (1.81, n = 1,333) and **Good** (1.26, n = 47,383). This pattern is consistent with expectations where policyholders with a poor claims history tend to make claims more frequently, contributing to greater portfolio volatility. All pairwise comparisons are statistically significant, confirming that bonus score is a strong and meaningful predictor of claim frequency.

## Policy Type

The violin plot below examines whether `policy_type` has a significant effect on `claim_frequency`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_cf,
  x    = policy_type,
  y    = claim_frequency,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Claim Frequency by Policy Type",
  xlab = "Policy Type",
  ylab = "Claim Frequency",
  palette = "Set2"
)
```

The Kruskal-Wallis test confirms a statistically significant difference in claim frequency across policy types (p = 3.28e-298). **Comprehensive without Excess** policies have the highest median claim frequency of **2.00** (n = 4,836), followed by **Comprehensive with Excess** (1.40, n = 16,470). In contrast, Third-Party based policies have notably lower claim frequencies — **Third-Party Combined** (1.17, n = 24,157), **Third-Party with Glass** (1.18, n = 3,868) and **Third-Party** (1.22, n = 384). This suggests that comprehensive policyholders tend to make claims more frequently, possibly because broader coverage lowers the threshold for filing a claim. All pairwise comparisons are statistically significant.

## Municipality Type

The violin plot below examines whether `municipality_type` has a significant effect on `claim_frequency`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_cf,
  x    = municipality_type,
  y    = claim_frequency,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Claim Frequency by Municipality Type",
  xlab = "Municipality Type",
  ylab = "Claim Frequency",
  palette = "Set2"
)
```

The Kruskal-Wallis test confirms a statistically significant difference in claim frequency across municipality types (p = 6.41e-05). However the differences are very small, **Inland** municipalities have the highest median claim frequency of **1.29** (n = 33,163), closely followed by **Coastal** (1.28, n = 13,502) and **Islands** (1.17, n = 3,050). The effect size is negligible, suggesting that while statistically significant, municipality type has minimal practical impact on claim frequency.

## Circulation Area

The violin plot below examines whether `circulation_area` has a significant effect on `claim_frequency`.

```{r}
#| warning: false
#| fig-height: 8
ggbetweenstats(
  data = insurance_cf,
  x    = circulation_area,
  y    = claim_frequency,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Claim Frequency by Circulation Area",
  xlab = "Circulation Area",
  ylab = "Claim Frequency",
  palette = "Set2"
)
```

The Mann-Whitney test confirms a statistically significant difference in claim frequency between circulation areas (p = 4.86e-13). **Urban** drivers have a higher median claim frequency of **1.33** (n = 28,271) compared to **Rural** drivers at **1.23** (n = 21,444). This aligns with expectations where urban environments with higher traffic density tend to result in more frequent claims. However, the effect size is small (r = -0.04), indicating that while the difference is statistically significant, circulation area has limited practical impact on portfolio volatility.

## Driver Age

The scatterplot below examines whether `driver_age` has a significant relationship with `claim_frequency`.

```{r}
#| warning: false
#| fig-height: 8
ggscatterstats(
  data = insurance_cf,
  x    = driver_age,
  y    = claim_frequency,
  type = "nonparametric",
  messages = FALSE,
  title = "Claim Frequency by Driver Age",
  xlab = "Driver Age (years)",
  ylab = "Claim Frequency",
  marginal = TRUE
)
```

The Spearman correlation test reveals a statistically significant but very weak negative relationship between driver age and claim frequency (ρ = -0.04, p = 3.07e-20). The nearly flat blue trend line confirms that older drivers tend to have a very slightly lower claim frequency, contrary to the common assumption that older drivers are higher risk. The effect size is negligible, suggesting that driver age alone is not a meaningful predictor of portfolio volatility in this dataset.
:::

## 6.3 Distribution Comparison across Key Groups

To complement the statistical analysis in Sections 6.1 and 6.2, ridgeline plots are used to compare the full distribution of `loss_ratio` and `claim_frequency` across key grouping variables. Ridgeline plots are particularly useful for visualising how distributions shift across multiple groups simultaneously.

::: panel-tabset
## Loss Ratio by Policy Type

```{r}
#| code-fold: true
#| warning: false
insurance_lr %>%
  ggplot(aes(x = loss_ratio,
             y = policy_type,
             fill = policy_type)) +
  geom_density_ridges(alpha = 0.7,
                      scale = 0.9,
                      quantile_lines = TRUE,
                      quantiles = 2) +
  scale_fill_manual(values = c(
    "Comprehensive with Excess"    = "#4a7c6f",
    "Comprehensive without Excess" = "#7aadab",
    "Third-Party"                  = "#d4e8e1",
    "Third-Party Combined"         = "#2d5a51",
    "Third-Party with Glass"       = "#a8d5c2")) +
  labs(title = "Distribution of Loss Ratio by Policy Type",
       subtitle = "Vertical line shows median",
       x = "Loss Ratio",
       y = "Policy Type") +
  theme_minimal() +
  theme(legend.position = "none")
```

The ridgeline plot provides a clearer view of how loss ratio distributions differ across policy types. **Comprehensive with Excess** and **Third-Party Combined** show the narrowest and most concentrated distributions, peaking sharply around 1–2, indicating consistent and predictable loss behaviour. **Comprehensive without Excess** and **Third-Party with Glass** display 
slightly wider distributions with longer right tails, suggesting greater variability. Most notably, **Third-Party** policies exhibit a distinctive bimodal distribution with two peaks, one around 1–2 and another around 4–5, suggesting two distinct subgroups with very different loss behaviours. This may be attributed to its small sample size (n = 241) where a few extreme cases disproportionately influence the distribution shape.

Overall, the ridgeline plot reinforces that Comprehensive policies generate more stable and predictable loss ratios, while Third-Party policies, particularly basic Third-Party exhibit greater variability and higher losses.


## Claim Frequency by Bonus Score

```{r}
#| code-fold: true
#| warning: false
insurance_cf %>%
  ggplot(aes(x = claim_frequency,
             y = bonus_score,
             fill = bonus_score)) +
  geom_density_ridges(alpha = 0.7,
                      scale = 0.9,
                      quantile_lines = TRUE,
                      quantiles = 2) +
  scale_fill_manual(values = c(
    "Good"    = "#4a7c6f",
    "Neutral" = "#7aadab",
    "Bad"     = "#d4e8e1")) +
  labs(title = "Distribution of Claim Frequency by Bonus Score",
       subtitle = "Vertical line shows median",
       x = "Claim Frequency",
       y = "Bonus Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

The ridgeline plot clearly shows distinct distribution shapes across bonus score groups. **Good** bonus score policies show a sharp, concentrated peak around 1, reflecting consistently low claim frequencies among the majority of the portfolio. **Neutral** policies display a wider bimodal distribution with two visible peaks, suggesting greater variability in claim behaviour.

**Bad** bonus score policies (median = 1.90) show the widest and most right-skewed distribution, with a long tail extending beyond 6, visually confirming the statistical findings from Section 6.2 that policyholders with a poor claims history not only make more frequent claims but also exhibit the most unpredictable claim behaviour in the portfolio.

:::


# 7 Summary

This study applied univariate and bivariate visual analytics techniques to explore a real-world motor vehicle insurance portfolio from a Spanish non-life insurance company, with the aim of uncovering key factors influencing portfolio profitability and volatility.

## Key Findings

**Portfolio Overview**

The portfolio is dominated by middle-aged drivers holding Third-Party Combined policies, with the vast majority of policyholders recording no claims. This suggests a generally low-risk customer base, with a small but significant minority of high-risk policyholders driving most of the portfolio's losses.

**Profitability (Loss Ratio)**

Policy type emerges as the strongest differentiator of portfolio profitability. Basic Third-Party policies are the least profitable despite being the least common, while Comprehensive with Excess policies generate the most consistent and stable loss ratios. This suggests that broader coverage does not necessarily translate to higher losses, in fact, comprehensive policyholders tend to be more careful drivers who pay higher premiums.

Fuel type and circulation area, while statistically significant, have negligible practical impact on profitability and are unlikely to be meaningful underwriting factors on their own.

**Volatility (Claim Frequency)**

Bonus score is the most powerful predictor of claim frequency. Policyholders with a Bad bonus score reflects a history of past claims, make significantly more frequent claims, contributing disproportionately to portfolio volatility. This confirms that past claims history is a reliable indicator of future claim behaviour.

Comprehensive without Excess policies stand out as the most volatile policy type, likely because the absence of an excess removes the financial disincentive to file small claims. In contrast, Third-Party Combined policies show the most stable and predictable claim behaviour.

## Recommendations for Future Research

- **Confirmatory Data Analysis** — Future studies should incorporate additional statistical tests such as regression modelling to validate the findings from this exploratory analysis and establish more definitive relationships between the identified risk factors and portfolio outcomes.

- **Exploring Additional Variables** — A more exhaustive analysis incorporating a wider range of variables could yield deeper insights into portfolio profitability and volatility. Variables such as driver occupation, annual mileage, claims history duration and vehicle usage type could provide a more complete picture of risk segmentation.

- **Longitudinal Analysis** — Conducting longitudinal studies to track changes in loss ratios and claim frequencies over a longer time period can reveal trends and seasonal patterns that a three-year snapshot may not capture.

- **Incorporating External Data** — As highlighted by the Swiss Re Institute (2019), enriching the portfolio data with external sources such as traffic density, weather patterns and regional accident statistics could further refine segmentation and improve underwriting strategy.


# 8 References

Swiss Re Institute. (2019). *Advanced analytics: Unlocking new frontiers in P&C insurance* (sigma No. 4/2019). Swiss Re Management Ltd. https://www.swissre.com/dam/jcr:3257ef4b-fe20-45e2-9700-7677702407af/sigma4_2019_en.pdf

Pesantez-Narvaez, J. (2023). *Motor vehicle insurance portfolio dataset*. Mendeley Data, V1. https://data.mendeley.com/datasets/sw4jmdb2sm/1