---
title: "Executive Summary: Motor Vehicle Insurance Loss Drivers"
subtitle: "Take-home Exercise 1"
author: "Carina Peh"
date: "May 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    transition: fade
    preview-links: true
    smaller: true
    self-contained: true
    footer: "ISSS608 Visual Analytics and Applications | Carina Peh"
    css: slides.css
execute:
  echo: false
  warning: false
  message: false
  freeze: auto
editor: visual
---

```{r}
#| label: setup
#| include: false
pacman::p_load(tidyverse, patchwork, scales, ggridges, ggdist, ggstatsplot, ggrepel)

insurance <- read_rds("insurance_clean.rds")

insurance_lr <- insurance %>%
  filter(loss_ratio > 0, loss_ratio < 10)

insurance_cf <- insurance %>%
  filter(claim_frequency > 0, claim_frequency < 10)

claim_summary <- insurance %>%
  mutate(claim_status = ifelse(has_claim == 1, "Has Claim", "No Claim")) %>%
  count(claim_status) %>%
  mutate(pct = n / sum(n) * 100)

policy_summary <- insurance %>%
  count(policy_type) %>%
  mutate(pct = n / sum(n) * 100)

ti_stats <- insurance %>%
  filter(total_incurred > 0, total_incurred < 10000) %>%
  summarise(mean_ti = mean(total_incurred), median_ti = median(total_incurred))

lr_stats <- insurance_lr %>%
  summarise(mean_lr = mean(loss_ratio), median_lr = median(loss_ratio))

cf_stats <- insurance_cf %>%
  summarise(mean_cf = mean(claim_frequency), median_cf = median(claim_frequency))
```

::: notes
This presentation summarises Take-home Exercise 1 in an executive format. The analysis uses visual analytics to uncover loss drivers in a Spanish motor vehicle insurance portfolio. Loss ratio measures profitability, claim frequency measures volatility, and has_claim distinguishes policies with and without claims.
:::

## Contents

1. Study Objective and Data Scope
2. Portfolio Profile
3. Claim Occurrence and Severity
4. Profitability Patterns
5. Volatility Patterns
6. Portfolio Loss Ratio Trend
7. Key Loss Drivers — Profitability
8. Key Loss Drivers — Volatility
9. Risk Concentration: Severity vs Frequency
10. Executive Implications
11. Recommendations and Next Steps

::: notes
The executive summary follows a decision-focused storyline. It begins with the purpose of the analysis and data scope, then moves into the portfolio profile and claim behaviour. The middle section interprets loss ratio and claim frequency distributions and examines the year-on-year trend. The final section highlights the most important loss drivers and translates them into practical recommendations for pricing, underwriting and future analysis.
:::

## Study Objective and Data Scope

**Objective:** identify visual and statistical patterns that explain poor profitability and high volatility.

| Area | Summary |
|---|---|
| Dataset | Spanish motor vehicle insurance portfolio |
| Original data | 354,140 policy records, 47 variables |
| Analytical data | 352,338 records after missing value treatment |
| Period | 2022 to 2024 |
| Key outcomes | Loss ratio, claim frequency, has claim |

- **Loss ratio** = total incurred / total premium — measures profitability
- **Claim frequency** = total claims / total exposure — measures volatility
- **Has claim** = 1 if claims > 0 — identifies claiming policies

::: notes
The analysis is based on a large real-world insurance portfolio. After cleaning, only 1,802 records were removed, so the final sample remains highly representative. The three key outcome measures answer two executive questions: which policies are less profitable based on loss ratio, and which segments are more volatile based on claim frequency.
:::

## Portfolio Profile

```{r}
#| fig-width: 8.5
#| fig-height: 3.5
p1 <- policy_summary %>%
  ggplot(aes(x = reorder(policy_type, -n), y = n)) +
  geom_col(fill = "#4a7c6f") +
  geom_text(aes(label = paste0(round(pct, 1), "%")), vjust = -0.35, size = 3) +
  scale_y_continuous(labels = comma) +
  labs(title = "Policy Type Distribution", x = NULL, y = "Number of policies") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1, size = 7))

p2 <- insurance %>%
  count(bonus_score) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = reorder(bonus_score, -n), y = n)) +
  geom_col(fill = "#7aadab") +
  geom_text(aes(label = paste0(round(pct, 1), "%")), vjust = -0.35, size = 3) +
  scale_y_continuous(labels = comma) +
  labs(title = "Bonus Score Distribution", x = NULL, y = "Number of policies") +
  theme_minimal()

p1 + p2
```

- **Third-Party Combined (58%)** dominates — performance issues here affect the whole portfolio
- **96.5%** of policyholders have a Good bonus score — portfolio appears low risk on the surface
- However, aggregate stability can mask concentrated pockets of high-risk segments

::: notes
The portfolio is heavily concentrated in Third-Party Combined policies at 58%, with Comprehensive with Excess second at 25.6%. Basic Third-Party is only 1.6%, patterns in this group must be interpreted with caution. The high proportion of Good bonus score policyholders suggests a surface-level low-risk profile, but visual analytics reveals that smaller high-risk segments can still drive most losses.
:::

## Claim Occurrence and Severity

```{r}
#| fig-width: 8.5
#| fig-height: 3.5
p3 <- claim_summary %>%
  ggplot(aes(x = claim_status, y = n, fill = claim_status)) +
  geom_col() +
  geom_text(aes(label = paste0(round(pct, 1), "%")), vjust = -0.5, size = 4) +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("Has Claim" = "#4a7c6f", "No Claim" = "#7aadab")) +
  labs(title = "Claim Occurrence", x = NULL, y = "Number of policies") +
  theme_minimal() +
  theme(legend.position = "none")

p4 <- insurance %>%
  filter(total_incurred > 0, total_incurred < 10000) %>%
  ggplot(aes(x = total_incurred)) +
  geom_histogram(binwidth = 200, fill = "#4a7c6f", colour = "white") +
  geom_vline(xintercept = ti_stats$mean_ti, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = ti_stats$median_ti, colour = "blue", linetype = "dashed") +
  scale_x_continuous(labels = comma) +
  labs(title = "Total Incurred Distribution", x = "Total incurred (€)", y = "Count") +
  theme_minimal()

p3 + p4
```

- **85.7%** of policies record no claims — but the **14.3%** with claims drive all losses
- Total incurred is **right-skewed** — most claims below €2,500 but occasional large losses remain
- **Mean (€1,271.98) > Median (€898.68)** — severe claims pull the average upward

::: notes
Most policies do not generate claims. The 85.7% no-claim rate is typical in motor insurance, but creates an analytical challenge, the majority group can make the portfolio look calm while the smaller claiming group drives most incurred costs. Total incurred is right-skewed with the mean substantially higher than the median, confirming that a small number of high-cost claims disproportionately affect the portfolio.
:::

## Profitability Patterns: Loss Ratio

```{r}
#| fig-width: 8.5
#| fig-height: 3.5
insurance_lr %>%
  ggplot(aes(x = loss_ratio)) +
  geom_histogram(binwidth = 0.1, fill = "#4a7c6f", colour = "white") +
  geom_vline(xintercept = lr_stats$mean_lr, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = lr_stats$median_lr, colour = "blue", linetype = "dashed") +
  annotate("text", x = lr_stats$mean_lr + 0.3, y = Inf, vjust = 2,
           label = paste("Mean:", round(lr_stats$mean_lr, 2)), colour = "red", size = 3) +
  annotate("text", x = lr_stats$median_lr + 0.3, y = Inf, vjust = 4,
           label = paste("Median:", round(lr_stats$median_lr, 2)), colour = "blue", size = 3) +
  labs(title = "Distribution of Loss Ratio",
       subtitle = "Policies with loss ratio between 0 and 10",
       x = "Loss ratio", y = "Count") +
  theme_minimal()
```

- Loss ratio is **highly right-skewed** — most policies record loss ratios between 1.0 and 2.5
- **Mean (2.58) > Median (2.0)** — extreme high-loss policies pull the average upward
- Long right tail indicates a minority of **catastrophic loss policies** driving most profitability risk

::: notes
Loss ratio is highly right-skewed among claiming policies. Many observations sit between 1.0 and 2.5, meaning incurred costs often exceed premiums for these policies. The long right tail indicates a small number of very high loss ratio cases. The mean is higher than the median, confirming the influence of extreme values. Profitability risk is concentrated rather than evenly spread across the portfolio.
:::

## Volatility Patterns: Claim Frequency

```{r}
#| fig-width: 8.5
#| fig-height: 3.5
insurance_cf %>%
  ggplot(aes(x = claim_frequency)) +
  geom_histogram(binwidth = 0.1, fill = "#4a7c6f", colour = "white") +
  geom_vline(xintercept = cf_stats$mean_cf, colour = "red", linetype = "dashed") +
  geom_vline(xintercept = cf_stats$median_cf, colour = "blue", linetype = "dashed") +
  annotate("text", x = cf_stats$mean_cf + 0.3, y = Inf, vjust = 2,
           label = paste("Mean:", round(cf_stats$mean_cf, 2)), colour = "red", size = 3) +
  annotate("text", x = cf_stats$median_cf + 0.3, y = Inf, vjust = 4,
           label = paste("Median:", round(cf_stats$median_cf, 2)), colour = "blue", size = 3) +
  labs(title = "Distribution of Claim Frequency",
       subtitle = "Policies with claim frequency between 0 and 10",
       x = "Claim frequency", y = "Count") +
  theme_minimal()
```

- Claim frequency is **right-skewed** — most claiming policies record around 1 claim per year
- **Mean (1.95) > Median (1.28)** — high-frequency claimants pull the average upward
- Portfolio volatility is driven by a **small group of frequent claimants**

::: notes
Claim frequency is also right-skewed. Most policies with claims are clustered around one claim per policy year. The mean claim frequency is 1.95, compared with a median of 1.28. This gap shows that high-frequency claimants pull the average upward. The distribution supports the decision to use non-parametric tests and highlights that portfolio volatility is mainly driven by a relatively small group of frequent claimants.
:::

## Portfolio Loss Ratio Trend

```{r}
#| fig-width: 8.5
#| fig-height: 3.5
insurance %>%
  group_by(year) %>%
  summarise(loss_ratio = sum(total_incurred) / sum(total_premium)) %>%
  ggplot(aes(x = year, group = 1)) +
  geom_line(colour = "#4a7c6f", linewidth = 1.2) +
  geom_point(colour = "#4a7c6f", size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  geom_text(aes(y = loss_ratio,
                label = round(loss_ratio, 3)),
            vjust = -0.8, size = 3) +
  labs(title = "Portfolio Loss Ratio by Year",
       subtitle = "Red dashed line = break-even (loss ratio = 1.0)",
       x = "Year", y = "Loss Ratio") +
  theme_minimal()
```

- Loss ratio rising steadily — **0.657 (2022) → 0.665 (2023) → 0.747 (2024)**
- Portfolio remains **below break-even** but the upward trend is a concern
- A **13.7% increase** over three years suggests claims costs are outpacing premiums
- Urgent to identify and address underlying loss drivers before the trend continues

::: notes
The line chart shows that while the portfolio remains profitable overall, the loss ratio has been deteriorating year on year. The jump from 0.665 to 0.747 between 2023 and 2024 is particularly notable — a single-year increase of 12.3%. If this rate of deterioration continues, the portfolio could approach the break-even point within two to three years. This trend provides the motivation for the deep-dive analysis into loss drivers that follows.
:::

## Key Loss Drivers: Profitability

```{r}
#| warning: false
#| fig-width: 8.5
#| fig-height: 3.8
ggbetweenstats(
  data = insurance_lr,
  x    = policy_type,
  y    = loss_ratio,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Loss Ratio by Policy Type",
  xlab = "Policy Type",
  ylab = "Loss Ratio",
  palette = "Set2"
)
```

- **Kruskal-Wallis p = 5.09e-92** — policy type significantly differentiates loss ratio
- **Basic Third-Party** has the highest median loss ratio (4.29) — least profitable
- **Comprehensive with Excess** is the most stable (median 1.74) — best profitability

::: notes
Policy type is the strongest profitability differentiator. The Kruskal-Wallis test confirms highly significant differences (p = 5.09e-92). Basic Third-Party has the highest median loss ratio at 4.29, though the small sample size (n = 241) warrants caution. Comprehensive with Excess is most stable at 1.74. The excess structure encourages policyholders to absorb minor costs themselves, resulting in lower and more predictable loss ratios for the insurer.
:::

## Key Loss Drivers: Volatility

```{r}
#| warning: false
#| fig-width: 8.5
#| fig-height: 3.8
ggbetweenstats(
  data = insurance_cf,
  x    = bonus_score,
  y    = claim_frequency,
  type = "nonparametric",
  mean.ci = TRUE,
  pairwise.comparisons = TRUE,
  pairwise.display = "significant",
  p.adjust.method = "fdr",
  messages = FALSE,
  title = "Claim Frequency by Bonus Score",
  xlab = "Bonus Score",
  ylab = "Claim Frequency",
  palette = "Set2"
)
```

- **Kruskal-Wallis p = 8.47e-61** — bonus score significantly differentiates claim frequency
- **Bad** bonus score has the highest median claim frequency (1.98) — most volatile
- Past claims history is a **reliable predictor** of future claim behaviour

::: notes
Bonus score is the strongest volatility differentiator. The Kruskal-Wallis test confirms highly significant differences (p = 8.47e-61). Bad bonus score policyholders have the highest median claim frequency at 1.98, followed by Neutral at 1.81 and Good at 1.26. This aligns with actuarial expectations — past claims history reliably predicts future claim behaviour. All pairwise comparisons are statistically significant.
:::

## Risk Concentration: Severity vs Frequency

```{r}
#| fig-width: 8.5
#| fig-height: 3.8
insurance %>%
  filter(has_claim == 1) %>%
  group_by(policy_type) %>%
  summarise(
    avg_severity  = mean(total_incurred),
    avg_frequency = mean(claim_frequency),
    total_premium = sum(total_premium)
  ) %>%
  ggplot(aes(x = avg_frequency,
             y = avg_severity,
             size = total_premium,
             colour = policy_type,
             label = policy_type)) +
  geom_point(alpha = 0.7) +
  geom_text_repel(size = 3, box.padding = 0.5) +
  scale_size_continuous(range = c(3, 15), labels = comma) +
  scale_x_continuous(expand = expansion(mult = 0.2)) +
  scale_y_continuous(expand = expansion(mult = 0.2)) +
  scale_colour_manual(values = c(
    "Comprehensive with Excess"    = "#4a7c6f",
    "Comprehensive without Excess" = "#7aadab",
    "Third-Party"                  = "#d4e8e1",
    "Third-Party Combined"         = "#2d5a51",
    "Third-Party with Glass"       = "#a8d5c2")) +
  labs(title = "Severity vs Frequency by Policy Type",
       subtitle = "Bubble size = total premium volume",
       x = "Average Claim Frequency",
       y = "Average Claim Severity (€)") +
  theme_minimal() +
  theme(legend.position = "none")
```

- **Comprehensive without Excess** — highest frequency AND severity, low premium — most loss-making
- **Third-Party Combined** — highest frequency, largest premium volume — frequency risk at scale
- **Comprehensive with Excess** — lower frequency, stable severity — healthiest segment

::: notes
The bubble chart combines three dimensions — frequency, severity and premium volume — in a single view. Comprehensive without Excess sits in the upper-right quadrant with both high frequency and high severity, yet generates relatively little premium. This is the most financially exposed segment. Third-Party Combined is the largest segment by premium but driven primarily by frequency risk. Comprehensive with Excess represents the most balanced risk-premium relationship in the portfolio.
:::

## Executive Implications

**The portfolio appears stable overall, but loss pressure is concentrated in selected segments.**

- **85.7%** of policies record no claims — but the claiming minority drives all losses
- Loss ratio rising **13.7% over three years** — profitability is deteriorating
- **Policy type** is the clearest profitability differentiator — Third-Party products are less profitable
- **Bonus score** is the clearest volatility differentiator — poor claims history predicts future frequency
- **Comprehensive without Excess** poses disproportionate risk — high frequency, high severity, low premium

::: notes
The main executive insight is that aggregate stability masks pockets of concentrated risk. The year-on-year trend shows the portfolio is moving in the wrong direction. Policy type and bonus score are the two most actionable risk factors. Comprehensive without Excess is particularly concerning as it combines the worst of both profitability and volatility dimensions. Immediate attention to pricing adequacy and underwriting criteria for these segments is warranted.
:::

## Recommendations and Next Steps

1. Review **pricing adequacy** for Third-Party based policies
2. Monitor **Bad and Neutral bonus score** groups more closely
3. Investigate **Comprehensive without Excess** for small claim filing behaviour
4. Apply **regression modelling** to confirm and quantify exploratory findings
5. Enrich portfolio with **traffic density, weather and regional accident data**

::: notes
The recommendations translate visual findings into action. Pricing adequacy should be reviewed for Third-Party policies due to weaker profitability. Bad and Neutral bonus score groups warrant closer monitoring given higher volatility. Comprehensive without Excess should be examined as the absence of an excess lowers the threshold for filing small claims. Future work should move from exploratory to confirmatory modelling and incorporate external data enrichment as recommended by the Swiss Re Institute (2019).
:::