---
title: "In-Class Exercise 6: Designing Web-enabled Visual Analytics Application with Shiny"
author: "Carina Peh"
date: "May 23, 2026"
date-modified: "last-modified"
description: "Building interactive web-enabled visual analytics applications using R Shiny, covering UI layout, server logic, and reactivity."
execute:
  eval: false
  echo: true
  warning: false
  freeze: true
---

## Overview

In this in-class exercise, we explored **Shiny**, R's framework for building interactive, web-enabled visual analytics applications. Unlike static ggplot2 visualisations, Shiny allows users to interact with data dynamically through widgets such as sliders and dropdowns — without needing to write any JavaScript or HTML.

## Reflections

Building the Shiny app helped me understand the clear separation between the **UI** (what the user sees) and the **server** (the logic behind what gets displayed). What clicked for me was how reactivity works — outputs don't just run once, they automatically re-execute whenever a relevant input changes. This is a fundamentally different mental model from writing a standard R script top to bottom.

I also noticed a practical difference between `aes()` and `aes_string()` — when the variable name comes from a reactive input (a string), `aes_string()` is needed so ggplot2 can interpret it correctly. Small details like this matter a lot when inputs are dynamic.

## Key Concepts

### Shiny Building Blocks

Every Shiny app has two components:

- **UI (User Interface)** — defines the layout and widgets the user interacts with
- **Server** — contains the R logic that generates outputs in response to inputs

### Reactivity

Shiny's reactivity system ensures outputs update automatically when inputs change. The key reactive functions are:

| Function | Purpose |
|---|---|
| `reactive()` | Creates a reactive expression that caches its result |
| `observe()` | Runs code for side effects when inputs change |
| `eventReactive()` | Only reacts when a specific event occurs (e.g. button click) |
| `isolate()` | Reads a reactive value without triggering re-execution |

## Pupils Exam Results Dashboard

### Loading Packages and Data

```{r}
pacman::p_load(shiny, tidyverse)
exam <- read_csv("Exam_data.csv")
```

### UI

The UI uses `sidebarLayout()` with two widgets — a `selectInput()` for subject selection and a `sliderInput()` for controlling the number of histogram bins.

```{r}
ui <- fluidPage(
  titlePanel("Pupils Exam Results Dashboard"),
  sidebarLayout(
    sidebarPanel(
      selectInput(inputId = "variable",
                  label = "Subject:",
                  choices = c("English" = "ENGLISH",
                              "Maths"   = "MATHS",
                              "Science" = "SCIENCE"),
                  selected = "ENGLISH"),
      sliderInput(inputId = "bins",
                  label = "Number of Bins",
                  min   = 5,
                  max   = 20,
                  value = 10)
    ),
    mainPanel(
      plotOutput("distPlot")
    )
  )
)
```

### Server

The server uses `aes_string()` to pass the reactive `input$variable` (a string) into ggplot2.

```{r}
server <- function(input, output) {
  output$distPlot <- renderPlot({
    ggplot(exam,
           aes_string(x = input$variable)) +
      geom_histogram(bins  = input$bins,
                     color = "black",
                     fill  = "light blue")
  })
}
```

### Running the App

```{r}
shinyApp(ui = ui, server = server)
```

## Live App

The deployed app is embedded below — try changing the subject and number of bins:

```{=html}
<iframe src="https://carinapeh.shinyapps.io/shiny_demo/" 
        width="100%" 
        height="600px"
        frameborder="0">
</iframe>
```